\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{enumitem}
\usepackage[colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}

\newcommand{\MSO}{\ensuremath{\mathrm{MSO}}}
\newcommand{\MSOtwo}{\ensuremath{\mathrm{MSO}_2}}
\newcommand{\Bags}{\ensuremath{\mathrm{BAGS}}}
\newcommand{\adh}{\ensuremath{\mathrm{adh}}}
\newcommand{\cone}{\ensuremath{\mathrm{cone}}}

\title{Mathematical Issues and Gaps in\\
  \texttt{lecture\_note\_expanded.tex}}
\author{}
\date{Review of the version dated June 19, 2026}

\begin{document}
\maketitle

\section*{Scope and summary}

This report reviews the mathematical argument in
\texttt{lecture\_note\_expanded.tex}.  It does not attempt a Lean design review.
Line numbers below refer to that source file.  The most serious findings in the
earlier draft were:
\begin{enumerate}[label=\textbf{C\arabic*.},leftmargin=*]
  \item the claimed reduction of \MSOtwo{} to ordinary graph \MSO{} on the
    uncoloured incidence graph is false;
  \item the definition of a legal tree omits the requirement that the root
    boundary be empty, so the decoding and inverse lemmas are false as stated;
  \item the cone-gluing identity is not an equality of the rooted, labelled
    graphs that were defined; and
  \item the running-time statement cannot hold for the stated input format,
    since it omits both the formula length and the size of the supplied
    decomposition.
\end{enumerate}
The remaining findings are mainly missing lemmas or underspecified standard
results.  They appear repairable, but several of them are essential links in
the proof rather than cosmetic details.

\section{Critical mathematical problems}

\subsection{The incidence-graph reduction does not define the two sorts}

\paragraph{Location.} The \MSOtwo{} corollary and its proof (the coloured
incidence graph \(\widehat I(G)\)).

The note proposes, in the uncoloured incidence graph \(I(G)\),
\[
  \operatorname{edg}(z) := \exists x\,\operatorname{inc}(x,z),
  \qquad
  \operatorname{vtx}(z) := \neg\operatorname{edg}(z),
\]
and then replaces incidence by ordinary adjacency.  After that replacement,
\(\operatorname{edg}(z)\) merely says that \(z\) has a neighbour.  It holds for
every nonisolated original vertex as well as for every edge-vertex.  For
example, if \(G\) consists of one edge and its two endpoints, then every one of
the three vertices of \(I(G)\) satisfies this formula, while none satisfies the
proposed vertex formula.

More conceptually, forgetting the predicates naming the two sides of the
incidence structure loses precisely the sort information needed to interpret
typed vertex and edge variables.  Consequently the displayed translation is
not truth-preserving, and the \MSOtwo{} corollary is not proved.

\paragraph{Recommended repair: retain the two sorts as unary colours.}
Define the \emph{coloured incidence structure} \(\widehat I(G)\) as follows:
\begin{itemize}[leftmargin=*]
  \item its universe is the disjoint union \(V(G)\sqcup E(G)\);
  \item \(\mathsf{Vert}(z)\) holds exactly for elements coming from \(V(G)\);
  \item \(\mathsf{EdgeObj}(z)\) holds exactly for elements coming from
    \(E(G)\); and
  \item \(\operatorname{Adj}_I(z,z')\) holds exactly when one element is a
    vertex, the other is an edge-object, and the original vertex is an endpoint
    of the original edge.
\end{itemize}
Thus \(\mathsf{Vert}\) and \(\mathsf{EdgeObj}\) are part of the vocabulary; they
are not supposed to be reconstructed from the uncoloured adjacency relation.

Translate a typed \MSOtwo{} formula \(\psi\) into ordinary one-sorted \MSO{} over
\[
  \tau_I=\{\operatorname{Adj}_I,\mathsf{Vert},\mathsf{EdgeObj}\}.
\]
The atomic incidence formula becomes
\[
  \operatorname{inc}(x,e)^\dagger
  :=\operatorname{Adj}_I(x,e).
\]
Membership and equality are left unchanged.  The sorts are enforced at every
typed quantifier.  For first-order quantifiers, use
\[
\begin{aligned}
  (\exists x\!:\!V\;\theta)^\dagger
    &:=\exists x\bigl(\mathsf{Vert}(x)\wedge\theta^\dagger\bigr),\\
  (\exists e\!:\!E\;\theta)^\dagger
    &:=\exists e\bigl(\mathsf{EdgeObj}(e)\wedge\theta^\dagger\bigr).
\end{aligned}
\]
For monadic second-order quantifiers, define
\[
\begin{aligned}
  \operatorname{VertSet}(X)&:=
    \forall z\,(z\in X\to\mathsf{Vert}(z)),\\
  \operatorname{EdgeSet}(F)&:=
    \forall z\,(z\in F\to\mathsf{EdgeObj}(z)),
\end{aligned}
\]
and use
\[
\begin{aligned}
  (\exists X\!\subseteq\!V\;\theta)^\dagger
    &:=\exists X\bigl(\operatorname{VertSet}(X)\wedge\theta^\dagger\bigr),\\
  (\exists F\!\subseteq\!E\;\theta)^\dagger
    &:=\exists F\bigl(\operatorname{EdgeSet}(F)\wedge\theta^\dagger\bigr).
\end{aligned}
\]
Universal quantifiers can be translated dually, using implication instead of
conjunction, or by \(\forall q\,\theta\equiv\neg\exists q\,\neg\theta\).  A
straightforward induction on \(\psi\) then proves
\[
  G\models_{\MSOtwo}\psi
  \quad\Longleftrightarrow\quad
  \widehat I(G)\models_{\MSO}\psi^\dagger.
\]

To apply the main theorem, its MSO-over-graphs statement must therefore be
slightly generalized from the vocabulary \(\{\operatorname{adj}\}\) to a fixed
finite vocabulary containing adjacency and finitely many unary predicates.
The proof changes only locally: a letter of \(\Sigma_\omega\) records, in
addition to the induced adjacency relation on a bag, which unary predicates
hold of each coloured bag element.  Compatibility of a child root with its
parent requires agreement on both adjacency and these unary predicates.  The
alphabet remains finite, and the defining-tuple translation gains atomic
clauses for the unary predicates.

Finally, the treewidth bound used in the note is valid for this coloured
incidence structure because unary predicates do not affect treewidth.  Given a
decomposition of \(G\), for each edge \(e=uv\) choose a bag containing \(u,v\)
and attach a leaf bag \(\{u,v,e\}\).  This covers the two incidence edges
\(ue\) and \(ve\), preserves connectedness of the bags containing each original
vertex, and places the edge-object \(e\) in exactly one bag.  The resulting
width is at most \(\max(\omega,2)\).  The separate size estimate needed for the
running time is discussed below.

If one insists on a theorem whose vocabulary is exactly
\(\{\operatorname{adj}\}\), the unary colours can instead be encoded by fixed
bounded-size graph gadgets.  That route requires an additional interpretation
theorem proving that the gadgets are logically recognizable and that they
increase treewidth only by a bounded amount.  Keeping the two unary predicates
is the simpler and more transparent repair.

\subsubsection*{How the rest of the proof should be adapted}

The remaining proof should not be read as a proof only for plain graph
structures \((V,\operatorname{adj})\).  It should first be upgraded to a proof
for structures over an arbitrary fixed finite relational vocabulary \(\tau\),
and then applied to the particular vocabulary
\[
  \tau_I=\{\operatorname{adj},\mathsf{Vert},\mathsf{EdgeObj}\}.
\]
This is the clean mathematical way to replace the old graph-only target
vocabulary
\[
  \{\operatorname{adj}\}
\]
by the incidence-structure vocabulary \(\tau_I\).  The upgrade is mostly local,
but it should be stated explicitly.

First, treewidth must be understood through the Gaifman graph of a structure.
The Gaifman graph of a \(\tau\)-structure has the same universe as the
structure, and two distinct elements are adjacent if they appear together in
some tuple of some relation of arity at least two.  Unary predicates, such as
\(\mathsf{Vert}\) and \(\mathsf{EdgeObj}\), do not add Gaifman edges.  Thus they
do not change treewidth.  For the coloured incidence structure, the Gaifman
graph is just the usual incidence graph.

Second, a bag letter should no longer be only a small rooted graph.  It should
be a small rooted \(\tau\)-structure.  For \(\tau_I\), this means that a letter
records
\begin{itemize}[leftmargin=*]
  \item which labels are present in the bag;
  \item which labels are in the root/adhesion;
  \item which labelled pairs satisfy \(\operatorname{adj}\);
  \item which labelled elements satisfy \(\mathsf{Vert}\); and
  \item which labelled elements satisfy \(\mathsf{EdgeObj}\).
\end{itemize}
Since \(\omega\) and \(\tau_I\) are fixed, there are still only finitely many
possible letters.  In the general \(\tau\)-structure version, the same letter
records the complete induced \(\tau\)-structure on the labelled bag.

Third, compatibility and gluing must include the predicate data.  If a child
root element is identified with a parent bag element of the same colour, then
the two copies must agree on all unary predicates.  In the incidence case, one
must not be allowed to glue a \(\mathsf{Vert}\)-element to an
\(\mathsf{EdgeObj}\)-element.  More generally, the two pieces must agree on the
induced \(\tau\)-structure on the common labelled root.  After gluing, each
relation of the glued structure is the union of the corresponding relations in
the pieces, modulo the quotient identification.  The final ``forget labels
outside the adhesion'' operation is still needed, exactly as in the graph-only
version.

Fourth, the decoding and re-encoding lemmas need the same additional coherence
statements for predicates that they need for edges.  For example, if two
representatives \((t,i)\) and \((s,i)\) denote the same decoded element, then
legality must imply that they have the same unary type:
\[
  \mathsf{Vert}(t,i)\Longleftrightarrow\mathsf{Vert}(s,i),
  \qquad
  \mathsf{EdgeObj}(t,i)\Longleftrightarrow\mathsf{EdgeObj}(s,i).
\]
This proves that the decoded predicates are well-defined on equivalence
classes.  The re-encoding proof must then show not only that the decoded
adjacency relation on each bag recovers the original letter, but also that the
decoded unary predicates on each bag recover the original unary predicate data.

Fifth, the formula-translation section gains atomic cases for the new relation
symbols.  Previously the main atomic case was the graph adjacency atom.  In the
incidence setting the atomic cases include
\[
  \operatorname{adj}(x,y),\qquad
  \mathsf{Vert}(x),\qquad
  \mathsf{EdgeObj}(x),
\]
besides equality and membership.  The unary cases are simpler than adjacency:
one only checks that the representative chosen for \(x\) carries the relevant
unary predicate in its top bag.  Quantifiers and Boolean connectives are handled
exactly as before.

Finally, the tree-automaton step is unchanged in substance.  The automaton
theorem applies to labelled trees over a finite alphabet.  Replacing graph
letters by finite \(\tau_I\)-structure letters merely changes the finite
alphabet and the compiled tree sentence.  Thus the high-level pipeline becomes
\[
  G\models_{\MSOtwo}\psi
  \Longleftrightarrow
  \widehat I(G)\models_{\MSO}\psi^\dagger
  \Longleftrightarrow
  \text{the decomposition encoding of }\widehat I(G)\text{ is accepted.}
\]
So the original proof architecture survives, but every occurrence of
``graph'', ``edge relation'', and ``rooted graph letter'' in the encoding part
should be replaced by the corresponding fixed-vocabulary structure notion.

\paragraph{Status (2026-06).} Resolved in \texttt{lecture\_note\_expanded.tex}.
The main theorem is now stated for \MSO{} over
\(\tau_{\mathcal P}=\{\operatorname{adj}\}\cup\mathcal P\), where
\(\mathcal P\) is a fixed finite set of unary predicates.  The alphabet letters
are rooted \(\tau_{\mathcal P}\)-graphs, so they record both adjacency and unary
predicate tags.  The legality condition now requires agreement of the full
induced \(\tau_{\mathcal P}\)-structure on a child root, not merely agreement of
the adjacency relation.  The formula translation has atomic clauses for each unary
predicate \(Q(x)\).  The \MSOtwo{} corollary now uses the coloured incidence
graph \(\widehat I(G)\) with unary predicates \(\mathsf{Vert}\) and
\(\mathsf{EdgeObj}\), and the typed \MSOtwo{} quantifiers are translated by
guarding them with these predicates.  The proof also records that unary
predicates do not affect treewidth and that bounded treewidth gives the needed
linear size bound for the incidence structure.

\subsection{A legal tree may have a nonempty root boundary}

\paragraph{Location.} The legality definition, \(\varphi_{\mathrm{legal}}\), and the
decoding lemmas.

Legality checks compatibility only across parent--child edges.  It imposes no
condition on the root letter.  However, the root adhesion of every encoded
decomposition is empty.  The proof of the decoding lemma simply asserts that the
root has empty external root because it has no parent, even though the
letter at the root may specify a nonempty root set.

There is an immediate counterexample: take a one-node tree whose letter is
\((H,R)\) with \(R\neq\varnothing\).  It is legal vacuously under the legality
definition,
but it cannot be the encoding of any decomposition, since the root adhesion of
an encoding is empty.  Re-encoding its proposed decoding changes \(R\) to the
empty set.  Thus the encoding--decoding inverse lemma is false as stated.

\paragraph{Repair.} Add the condition \(R=\varnothing\) at the root to the
definition of legality and to \(\varphi_{\mathrm{legal}}\).  The finite formula
can require that every root carry a letter in
\(\{(H,R):R=\varnothing\}\).

\paragraph{Status (2026-06).} Resolved in \texttt{lecture\_note\_expanded.tex}.
The legality definition now requires \(R_r=\varnothing\) at the root,
\(\varphi_{\mathrm{legal}}\) gains a conjunct forcing every root node (a node
with no parent) to carry a letter \((H,R)\) with \(R=\varnothing\), and the proof
of the decoding lemma now cites this root condition instead of ``the root has no
parent''.

\subsection{The cone-gluing equation needs a final label restriction}

\paragraph{Location.} The rooted-graph and cone-graph definitions and the
cone-gluing lemma.

There are two different subsets in the definition of a rooted graph, and they
should not be conflated.  The root or boundary is \(R\), while the domain of the
partial labelling map \(\iota\) is only required to \emph{contain} \(R\); labels
outside the root are explicitly permitted.  In the expanded note the two
relevant definitions are
\[
\begin{aligned}
  \text{node graph at }t
    &=\bigl(G[\beta(t)],\adh(t),c|_{\beta(t)}\bigr),\\
  \text{cone graph at }t
    &=\bigl(G[\cone(t)],\adh(t),c|_{\adh(t)}\bigr).
\end{aligned}
\]
Thus the node graph really is temporarily labelled on all of \(\beta(t)\), even
though its distinguished boundary is only \(\adh(t)\).  By contrast, the cone
graph keeps labels only on the adhesion through which the completed cone will
later be attached to its parent.

The larger temporary label domain is necessary.  For a child \(t_i\), its
adhesion \(\adh(t_i)\) is contained in \(\beta(t)\), but need not be contained in
\(\adh(t)\).  To glue the child cone to the node graph at \(t\), vertices of
\(\adh(t_i)\) must therefore have labels available in the first operand.  For
example, it is possible that
\[
  \beta(t)=\{a,b\},\qquad \adh(t)=\{a\},\qquad
  \adh(t_i)=\{b\}.
\]
The label of \(b\) is needed while attaching the child, although it must be
forgotten after the entire cone at \(t\) has been constructed.

Now apply the definition of gluing: the root and the whole labelling map of the
result are inherited from the first argument.  Hence the right-hand side of
\[
  (H_t,R_t,\iota_t)\oplus(G_1,R_1,\iota_1)
    \oplus(G_2,R_2,\iota_2)
\]
retains the temporary labels on every vertex of \(\beta(t)\), but the left-hand
cone graph has labels only on \(R_t=\adh(t)\).  The underlying glued graph and
the root are the intended ones, but the partial labelling maps have different
domains.  Consequently the displayed equality in the cone-gluing lemma is not
literally an equality of the triples defined in the rooted-graph definition.

There is also a structural equality issue: disjoint union followed by quotient
identifications normally produces a graph canonically isomorphic to the cone
graph, not literally equal to it.

\paragraph{Repair.} After both gluings, explicitly forget every label outside
\(R_t\), just as the decoding construction later does.  If
\(\operatorname{forget}_{R_t}\) denotes restriction of the partial labelling
map to \(R_t\), the corrected statement is, up to the chosen quotient model,
\[
  (G_t,R_t,c|_{R_t})\cong
  \operatorname{forget}_{R_t}\Bigl(
    (H_t,R_t,c|_{\beta(t)})\oplus
    (G_1,R_1,c|_{R_1})\oplus
    (G_2,R_2,c|_{R_2})
  \Bigr).
\]
Here \(\cong\) is a label-preserving rooted-graph isomorphism.  Alternatively,
one may define a canonical quotient construction and use literal equality.

If the intended convention were instead that only root vertices may ever be
labelled, then the node graph would have labelling \(c|_{\adh(t)}\).  That does
not repair the proof: in the example above the child cannot be glued because
the first operand has no label for \(b\).  The temporary labels on the whole
parent bag, followed by restriction to the parent adhesion, are the mechanism
that makes the recursive gluing work.

\paragraph{Status (2026-06).} Resolved in \texttt{lecture\_note\_expanded.tex}
by a different route than the one recommended above.  Rather than keeping the
cone labelling \(c|_{\adh(t)}\) and inserting an explicit
\(\operatorname{forget}_{R_t}\) on the right-hand side, the cone graph at \(t\)
is redefined with labelling \(c|_{\beta(t)}\).  Since gluing inherits the
labelling of its first operand (the node graph, already labelled
\(c|_{\beta(t)}\)), the two sides of the cone-gluing identity then carry the same
labelling and the equality holds literally.  The decoding keeps its own
forgetting step, which stays consistent because re-encoding goes through node
graphs, not cone graphs.  The structural \(\cong\)-versus-\(=\) point is
acknowledged in the cone-gluing proof, which states the identity up to the
canonical label-preserving isomorphism of the disjoint-union-and-quotient model.

\subsection{The stated running time ignores part of the input}

\paragraph{Location.} Theorem~1 (Courcelle's theorem) and its proof.

The theorem takes both a formula \(\varphi\) and a decomposition \(T\) as input,
but claims time
\[
  f(\omega,\operatorname{qr}(\varphi))\,|V(G)|.
\]
This cannot be the total running time for the stated input representation.

First, formulas of fixed quantifier rank can have arbitrarily large syntactic
length, and an algorithm must at least read the input formula.  The proof only
shows that the translated sentence and automaton depend on \(\omega\) and on the
whole formula; it does not justify replacing dependence on \(\varphi\) by
dependence only on its quantifier rank.

Second, a supplied decomposition can have arbitrarily many redundant or
duplicate nodes even when \(G\) is fixed.  Constructing its labelled encoding
takes \(\Theta(|N(T)|)\) time, as the proof itself observes.  A later appeal to a
``usual linear-size normalization'' cannot remove the cost of reading an
arbitrarily large input decomposition.

\paragraph{Repair.} A standard precise statement is either
\[
  f(\omega,|\varphi|)\bigl(|V(G)|+|N(T)|\bigr)
\]
up to the chosen input encoding, or a theorem in which \(\varphi\) is fixed and
the supplied decomposition is assumed to be in a normalized form with
\(|N(T)|=O(|V(G)|)\).  If quantifier rank is intentionally used as the logical
parameter, the compilation/preprocessing cost for reading \(\varphi\) must be
stated separately and a quantifier-rank bound for the automaton construction
must be proved.

\paragraph{Status (2026-06).} Resolved in \texttt{lecture\_note\_expanded.tex}.
Theorem~1 now states the honest varying-formula, supplied-decomposition bound
\[
  f_{\mathcal P}(\omega,|\varphi|)\bigl(|V(G)|+|N(T)|\bigr),
\]
for a fixed finite unary vocabulary \(\mathcal P\).  The surrounding remark
explains how this specializes to the usual linear-time statement when
\(\varphi\) is fixed and the decomposition has been normalized so that
\(|N(T)|=O(|V(G)|)\).

\section{Substantial proof gaps}

\subsection{The bag-injective colouring proof is not valid as written}

\paragraph{Location.} The Bag-injective coloring fact and its proof.

The greedy paragraph chooses an arbitrary bag containing the current vertex
and counts previously coloured vertices only in that bag.  It does not show
that all previously coloured vertices whose bag-subtrees intersect the current
vertex's bag-subtree occur together in the chosen bag.  Thus the asserted
greedy step is not established.  The phrase ``intersection graph of the bags''
is also inaccurate: the relevant graph has one vertex for each graph vertex,
and two such vertices conflict when their connected bag-subtrees intersect.

The existence claim is true.  A complete proof can use the facts that
intersection graphs of subtrees of a tree are chordal, subtrees of a tree have
the Helly property, and hence every clique of the conflict graph occurs in one
bag and has size at most \(\omega+1\).  Perfectness of chordal graphs then gives
the required colouring.  For the algorithmic theorem, one must additionally
give an effective ordering/colouring procedure rather than merely an existence
argument.

\paragraph{Status (2026-06).} Resolved in \texttt{lecture\_note\_expanded.tex}
by a constructive greedy colouring rather than the chordal/perfect-graph
existence route.  The bag set \(\Bags(v)=\{t:v\in\beta(t)\}\) and the topmost node
\(\mathrm{top}(\Bags(v))\) of a connected node set are now introduced before the
colouring fact (and reused throughout the encoding section); the vertices are
coloured in order of nondecreasing depth of \(\mathrm{top}(\Bags(v))\) (a single
root-to-leaf sweep), each \(v\) receiving a colour unused on the already-coloured
vertices of its top bag.  The proof now establishes the key claim that, at the
moment \(v\) is coloured, every already-coloured vertex sharing a bag with \(v\)
lies in that top bag; since the bag has at most \(\omega+1\) vertices, a free
colour always exists.  This simultaneously fixes the invalid ``arbitrary bag''
greedy step and the inaccurate ``intersection graph of the bags'' phrasing, and
it yields the effective (linear-sweep) procedure required by the algorithmic
theorem.

\subsection{Cone gluing needs intersection and edge-localisation lemmas}

\paragraph{Location.} The cone-gluing lemma and its proof.

The proof asserts without proof that gluing identifies exactly all shared
vertices of the three pieces and that every edge of \(G[\cone(t)]\) occurs in a
bag in the subtree rooted at \(t\).  Both facts follow from the running-
intersection axiom, but neither follows merely from the definitions.

A complete proof should establish at least
\[
\begin{aligned}
  \cone(t_i)\cap\beta(t)&=\adh(t_i),\\
  \cone(t_1)\cap\cone(t_2)&\subseteq\beta(t),
\end{aligned}
\]
and then show that any edge whose endpoints lie in \(\cone(t)\) is recorded in
the parent bag or in one child cone.  The latter requires following the
connected occurrence subtrees of both endpoints to a bag covering the edge.
These lemmas are also needed later for decoding correctness.

\paragraph{Status (2026-06).} Resolved in \texttt{lecture\_note\_expanded.tex}.
A numbered \emph{Cone structure} lemma is now stated and proved just before the
cone-gluing lemma, with three parts: (i) \(\cone(s)\cap\beta(t)=\adh(s)\) for a
child \(s\) of \(t\); (ii) \(\cone(t_1)\cap\cone(t_2)\subseteq\adh(t_1)\cap\adh(t_2)\)
for distinct children; and (iii) every edge of \(G[\cone(t)]\) lies in a single bag
of the subtree at \(t\), hence in one of the three pieces \(G[\beta(t)]\),
\(G[\cone(t_1)]\), \(G[\cone(t_2)]\).  Each part is proved from the connectedness
of the occurrence sets \(\Bags(\cdot)\) (the running-intersection axiom) by routing
tree paths through \(s\), through the parent \(t\), or out of the subtree at \(t\).
The cone-gluing proof now cites parts (i)--(ii) for ``the quotient identifies
precisely the shared vertices'' and part (iii) for ``every edge appears in one of
the three pieces''.  These lemmas are reused for the decoding correctness items
below.

\subsection{The decoding needs a precise quotient and a proof of connectedness}

\paragraph{Location.} The decoding construction and the decoding
well-definedness lemmas.

Phrases such as ``a vertex represented by \(v(t,i)\)'' presuppose an equivalence
relation generated by the gluing identifications, but that relation is not
defined.  Consequently it is not yet precise when two representatives denote
the same decoded vertex, or why a representative has a well-defined colour.

The connectedness argument in the decoding well-definedness lemma also skips its
central step.  It
claims that a representative can be identified with one outside a selected
component only through \(R_t\), without proving this from the recursive gluings.
A clean repair is to define decoded vertices as equivalence classes of pairs
\((t,i)\), where identifications are generated across a parent--child edge
exactly for colours in the child root.  One then proves that each equivalence
class has a connected set of tree nodes, either directly from paths in the tree
or by induction through the bottom-up gluing.  This simultaneously proves the
running-intersection axiom for the decoded bags.

\paragraph{Status (2026-06).} Resolved in \texttt{lecture\_note\_expanded.tex}
along the recommended route.  The decoded graph is now defined as an explicit
quotient: on the \emph{slots} \(v(t,i)\) (\(i\in V(H_t)\)) the relation \(\sim\)
is generated by \(v(x,i)\sim v(p(x),i)\) for \(i\in R_x\) (child-root colours), a
decoded vertex is a \(\sim\)-class, and colour, bags, adjacency and predicates are
read off the slots.  Colour is well-defined because the generators preserve the
colour coordinate, and the predicate tags are well-defined because legality forces
\(H_x[R_x]=H_{p(x)}[R_x]\).  For connectedness, a new lemma introduces, for each
colour \(i\), the tree-edge subgraph \(\Gamma_i\) (edges \(\{x,p(x)\}\) with
\(i\in R_x\)) and proves the path characterization: \(v(s,i)\sim v(t,i)\) iff every
edge of the \(s\)--\(t\) tree path lies in \(\Gamma_i\).  Each decoded vertex's
occurrence set is then a component of \(\Gamma_i\); since \(\Gamma_i\)'s edges are
tree edges, the component is a subtree, so the occurrence set is connected and the
running-intersection axiom holds.  The old ``minimum tree distance / topmost node''
argument and the informal bottom-up ``forget labels'' decoding (with its
well-definedness lemma) are removed.  The path characterization also discharges
item~(a) of the re-encoding gaps below.

\subsection{The re-encoding argument omits decisive coherence lemmas}

\paragraph{Location.} The legal-iff-encoding lemma (the re-encoding direction).

The proof asserts that the decoded node graph is exactly the original letter.
That does not follow immediately from the definition of the decoded graph,
which is the union of edges contributed at all tree nodes.  One must rule out
the possibility that an edge contributed elsewhere becomes an additional edge
between two vertices of the decoded bag at \(t\).  Local legality should imply
the needed edge coherence, but that implication is not proved.

Likewise, one must prove that the intersection of the decoded child and parent
bags is exactly the root set recorded by the child letter.  Merely knowing that
the recorded root is contained in both bags gives only one inclusion.

The missing statements can be organized as follows:
\begin{enumerate}[label=(\alph*),leftmargin=*]
  \item two representatives \((s,i)\) and \((t,i)\) are equivalent exactly when
    the unique path between \(s\) and \(t\) carries colour \(i\) through every
    intervening child root;
  \item for a child \(x\) of \(y\),
    \(\beta(x)\cap\beta(y)\) is precisely the set represented by \(R_x\);
  \item the subgraph of the final decoded graph induced on \(\beta(t)\) is
    exactly \(H_t\).
\end{enumerate}
With these lemmas and the empty-root correction, encoding after decoding really
does return the original labelled tree.

\paragraph{Status (2026-06).} Resolved in \texttt{lecture\_note\_expanded.tex},
using the slot-quotient decoding.  Item~(a) is the path-characterization lemma
already added for the decoding (\(v(s,i)\sim v(t,i)\) iff the \(s\)--\(t\) tree
path lies in \(\Gamma_i\)).  Item~(b) is a new lemma: for a child \(x\) with
parent \(y\), \(\beta(x)\cap\beta(y)=\{[v(x,i)]:i\in R_x\}\); the nontrivial
inclusion uses (a) on the single edge \(\{x,y\}\).  Item~(c) is a new lemma: for
each node \(t\), the map \(i\mapsto[v(t,i)]\) is a \(\tau_{\mathcal P}\)-graph
isomorphism from the letter \(H_t\) onto the induced subgraph \(G[\beta(t)]\); the
``no spurious edge'' direction rules out a foreign edge by transporting it along
the \(u\)--\(t\) path, where (a) gives \(i,j\in R_z\) at every edge and legality
\(H_z[R_z]=H_{p(z)}[R_z]\) keeps \(ij\in E(H_\cdot)\) constant.  The re-encoding
direction of the encoding--decoding inverse lemma now cites (b) for the recorded
root and (c) for the recorded graph, so re-encoding returns
\((H_t,R_t)=\rho(t)\) at every node.

\subsection{The formula-translation correctness proof is only announced}

\paragraph{Location.} The formula-translation section and its correctness
paragraph.

The note says that a structural induction gives correctness, but does not state
or prove the induction invariant.  In particular, the reverse direction of an
existential-quantifier step uses the fact that every tuple satisfying the
vertex or set guard decodes to a graph object, while the forward direction uses
surjectivity of the defining-tuple representation.  Those are the main logical
steps and should be written out.

For a syntactic translation, all auxiliary variables in the formulas for
connectedness, top nodes, and defining tuples must also be chosen fresh so that
substitution does not capture variables.  This is usually handled silently by
working modulo alpha-renaming, but the convention should be stated.  The
correctness theorem should quantify over graph assignments and their
corresponding tuple assignments, and then prove the atomic, Boolean, individual-
quantifier, and set-quantifier cases separately.

\paragraph{Status (2026-06).} Resolved in \texttt{lecture\_note\_expanded.tex}.
The announced paragraph is replaced by a numbered \emph{Translation correctness}
lemma with the assignment-level invariant: for the encoding \((T,\rho)\) of
\((G,c,(T,\beta))\) and any graph assignment \(\alpha\) (individual variables to
vertices, set variables to vertex sets), with the corresponding tree assignment
\(\alpha^\star\) sending each tuple to the defining tuple of \(\alpha\)'s value,
\(G\models\theta[\alpha]\) iff \((T,\rho)\models\theta^\star[\alpha^\star]\).  The
proof is by induction with the atomic, Boolean, individual-quantifier, and
set-quantifier cases separated.  The existential vertex case is written out in
both directions: forward uses that the witness vertex's defining tuple satisfies
the vertex guard; reverse uses that every tuple satisfying the guard is the
defining tuple of a vertex (the recognition lemmas for the guards), and similarly
for set quantifiers via the set guard.  A fresh-variable convention is stated:
auxiliary bound variables in the guard and tree macros are renamed disjoint from
the tuple variables, with formulas read modulo \(\alpha\)-equivalence.  The
sentence-level equivalence \(G\models\varphi\) iff
\((T,\rho)\models\varphi_{\mathrm{tree}}\) is the empty-assignment corollary, and
the proof of Theorem~1 now cites this lemma.

\subsection{The exact effective tree-automata theorem is not stated}

\paragraph{Location.} The proof of Theorem~1 (the tree-automaton step).

The equivalence between regularity and \MSO{} definability is cited only as an
existence theorem.  To obtain an algorithm, the proof needs the effective
direction: from the particular tree \MSO{} sentence and finite alphabet, one
can compute a finite bottom-up automaton.  It should also specify whether the
trees are full binary ranked trees, ordered trees of outdegree at most two, or
trees padded by null leaves.  The automaton model and the logical vocabulary
must match this choice.

This is a legitimate external theorem, but the exact version required by the
proof should be isolated as a theorem with a citation.  If deterministic
automata are desired, include the standard determinisation step.

\paragraph{Status (2026-06).} Resolved in \texttt{lecture\_note\_expanded.tex}.
The effective direction is now isolated as a cited theorem (Thatcher--Wright,
Doner): from an \MSO{} sentence over the tree vocabulary
\(\{\operatorname{child}_1,\operatorname{child}_2\}\cup\{P_a\}\) one computes a
deterministic bottom-up finite tree automaton accepting exactly the
\(\Sigma\)-trees that satisfy it, with a run linear in the number of nodes.  The
tree/automaton model is fixed: \(\Sigma\)-trees are encoded as full binary ranked
trees by padding absent children with a null symbol \(\bot\), and the automaton is
determinised by the bottom-up subset construction.  The proof of Theorem~1 now
invokes this theorem instead of an existence statement, and the bibliography gains
the Thatcher--Wright, Doner, and TATA references.

\section{Definitions and boundary cases requiring clarification}

\subsection{Unary nodes are omitted from the recursive construction}

The input is called a binary tree-decomposition, which often means that every
node has at most two children.  Nevertheless, the cone-gluing lemma and decoding
recursion discuss only leaves and nodes with two children.  A node with exactly
one child is not covered.  Either ``binary'' must mean full binary and a
normalisation to that form must be given, or the one-child gluing case must be
added.  This choice must agree with the tree-automaton model.

\paragraph{Status (2026-06).} Resolved in \texttt{lecture\_note\_expanded.tex}.
``Binary'' is now defined as ``each node has at most two ordered children'' in the
\(\Sigma\)-tree definition.  The slot-quotient decoding adopted for the
encoding--decoding inverse uses only the parent--child relation, so leaves and
one- and two-child nodes are handled uniformly, with no arity case split.  The
cone-gluing lemma is restated for children \(t_1,\ldots,t_d\) (with \(d=0\) at a
leaf and \(d\le 2\) in the binary case), and the cone-structure lemma's
edge-localisation part is generalised to the same form.

\subsection{The ambient class of tree structures is implicit}

The formulas use uniqueness of parents, a unique root, acyclicity, and exactly
one letter per node.  These are valid because the text informally evaluates
only on labelled rooted trees.  If satisfaction is instead taken over arbitrary
structures for the vocabulary
\(\{\operatorname{child}_1,\operatorname{child}_2,P_a\}\), then tree axioms and
the exactly-one-letter axiom are missing.  The cleanest mathematical statement
is to define a \(\Sigma\)-tree as a finite rooted ordered tree together with a
labelling function \(\rho:N(T)\to\Sigma\), and restrict all logical languages and
automata to that class.

\paragraph{Status (2026-06).} Resolved in \texttt{lecture\_note\_expanded.tex}.
A definition now fixes the ambient class exactly as recommended: a
\(\Sigma\)-tree is a finite rooted ordered tree (each node with an optional first
and second child) together with a labelling \(\rho:N(T)\to\Sigma\), and all tree
\MSO{} formulas and automata are interpreted over this class.  Parent- and
root-uniqueness, acyclicity, and the exactly-one-label condition therefore hold by
definition rather than as axioms.

\subsection{Rooted-graph notation and equality need one convention}

The rooted-graph definition uses labels in \([k]\), while the later colour set is
\(\{0,\ldots,\omega\}\); it also introduces \(\oplus_k\) and later drops the
subscript.  These are harmless only after an explicit identification of the
label types.  More importantly, all gluing statements should consistently use
either literal equality of a chosen quotient construction or isomorphism
classes of rooted labelled graphs.

\paragraph{Status (2026-06).} Resolved in \texttt{lecture\_note\_expanded.tex}.
The colour/label convention is now fixed uniformly by
\[
  C_k=\{0\}\cup[k],
\]
with \([0]=\varnothing\), so the width-\(\omega\) colour set is
\(C_\omega=\{0,\ldots,\omega\}\).  A \(k\)-rooted
\(\tau_{\mathcal P}\)-graph now uses labels in \(C_k\), and the text explicitly
says that \(\oplus\) abbreviates \(\oplus_k\) when \(k\) is clear.  The note also
defines rooted label-preserving isomorphism and states the cone-gluing lemma
using \(\cong\), avoiding a commitment to a particular quotient model for
disjoint-union gluing.

\subsection{The incidence construction also needs a size bound}

Even after repairing the two sorts, \(I(G)\) has \(|V(G)|+|E(G)|\) vertices and
the displayed decomposition adds one bag per edge.  To retain a linear bound in
\(|V(G)|\), the proof should invoke the standard fact that a simple graph of
treewidth at most \(\omega\) has \(O(\omega|V(G)|)\) edges (for example via
\(\omega\)-degeneracy).  For fixed \(\omega\) this gives the required linear
size.  Without that sentence, the runtime conclusion does not follow from the
incidence construction alone.

\paragraph{Status (2026-06).} Resolved in \texttt{lecture\_note\_expanded.tex}.
The \MSOtwo{} corollary proof now cites a new lemma proving that a simple graph
with a tree-decomposition of width at most \(\omega\) has at most
\(\omega|V(G)|\) edges, so the coloured incidence graph has \(O_\omega(|V(G)|)\)
vertices.  The lemma is proved from scratch: at a leaf bag of a width-\(\omega\)
decomposition there is a vertex lying in a single bag, hence of degree at most
\(\omega\) (so \(G\) is \(\omega\)-degenerate), and an induction on \(|V(G)|\)
that removes such a vertex gives the edge bound.

\section{Suggested order of repair}

The dependencies suggest the following order:
\begin{enumerate}[leftmargin=*]
  \item correct the main complexity statement and decide the exact binary-tree
    representation;
  \item replace the \MSOtwo{} incidence argument by a coloured incidence
    structure (or a proved gadget reduction);
  \item repair legality by enforcing an empty root boundary;
  \item define gluing and decoded vertices precisely, preferably using an
    explicit equivalence relation on representatives;
  \item prove the path/equivalence, adhesion, edge-coherence, and cone-union
    lemmas before restating the encoding--decoding inverse theorem;
  \item replace the colouring paragraph by a complete subtree-intersection
    argument and, for the algorithmic theorem, an effective colouring method;
  \item state and prove the assignment-level translation theorem; and
  \item state the precise effective \MSO{}-to-tree-automaton theorem being used.
\end{enumerate}

After these repairs, the defining-pair and defining-tuple part of the note
(the defining-pair and defining-tuple lemmas) appears mathematically sound; its main dependency is a
fully proved encoding--decoding theorem.

\end{document}
