\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{enumitem}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem*{remark}{Remark}

\newcommand{\defeq}{\coloneqq}
\newcommand{\Nodes}{N}
\newcommand{\width}{\operatorname{width}}
\newcommand{\children}{\operatorname{children}}
\newcommand{\Occ}{\operatorname{Occ}}

\title{Nice Tree-Decomposition Normalization}
\author{}
\date{}

\begin{document}
\maketitle

\section{Statement}

Graphs are finite simple graphs.  A tree-decomposition of a graph \(G\) is a
pair \((T,\beta)\), where \(T\) is a finite nonempty tree and
\(\beta:\Nodes(T)\to 2^{V(G)}\) assigns a bag to each node, such that:
\begin{enumerate}[label=\textup{(TD\arabic*)},leftmargin=*]
  \item for every \(v\in V(G)\), some node \(t\in\Nodes(T)\) satisfies
    \(v\in\beta(t)\);
  \item for every edge \(uv\in E(G)\), some node \(t\in\Nodes(T)\) satisfies
    \(u,v\in\beta(t)\);
  \item for every \(v\in V(G)\), the set
    \[
      \Occ_T(v)\defeq\{t\in\Nodes(T):v\in\beta(t)\}
    \]
    is connected in \(T\).
\end{enumerate}
The width is at most \(\omega\) if every bag has size at most \(\omega+1\).

\begin{definition}[Nice tree-decomposition]
A rooted tree-decomposition \((T,\beta)\) is \emph{nice} if the following hold.
\begin{enumerate}[label=\textup{(N\arabic*)},leftmargin=*]
  \item The root bag is empty, and every leaf bag is empty.
  \item Every node has at most two children.
  \item If a node \(x\) has exactly one child \(y\), then exactly one vertex is
    changed across the edge \(xy\): either
    \[
      \beta(x)=\beta(y)\cup\{v\}\quad\text{for some }v\notin\beta(y),
    \]
    or
    \[
      \beta(y)=\beta(x)\cup\{v\}\quad\text{for some }v\notin\beta(x).
    \]
    In the first case \(x\) is an introduce node, and in the second case
    \(x\) is a forget node, relative to the top-down orientation.
  \item If a node \(x\) has exactly two children \(y_1,y_2\), then
    \[
      \beta(x)=\beta(y_1)=\beta(y_2).
    \]
    Such a node is a join node.
\end{enumerate}
\end{definition}

In the algorithmic statement, the input decomposition is represented by
adjacency lists for \(T\) and by lists for the bags.  We write
\(n=|\Nodes(T)|\).  The intended Courcelle setting fixes \(\omega\), so
\(O_\omega(n)\) means linear time with a constant depending on \(\omega\).  If
one writes every output bag explicitly and lets \(\omega\) vary, the same
algorithm has the uniform bound \(O(\omega^2 n)\); with a compact
one-vertex-update representation of path bags, the node-generation work is
\(O(\omega n)\).

\begin{theorem}[Algorithmic nice normalization]\label{thm:nice-normalization}
Let \(G\) be a finite simple graph.  Let \((T,\beta)\) be a
tree-decomposition of \(G\), and suppose that \(|\beta(t)|\le \omega+1\) for
every \(t\in\Nodes(T)\).  Write \(n=|\Nodes(T)|\).  There is an algorithm which,
given \((T,\beta)\), constructs a nice tree-decomposition
\((T^\ast,\beta^\ast)\) of \(G\) such that
\[
  |\beta^\ast(x)|\le \omega+1
  \qquad\text{for every }x\in\Nodes(T^\ast),
\]
and
\[
  |\Nodes(T^\ast)|\le (3\omega+5)n.
\]
The algorithm runs in time \(O(\omega n)\), and hence in time \(O(n)\) for
The algorithm runs in time \(O_\omega(n)\), hence in time \(O(n)\) for fixed
\(\omega\).  More explicitly, with explicit bag-list output it runs in
\(O(\omega^2 n)\) time.  In particular, if the original width is exactly \(w\),
then taking \(\omega=w\) gives a nice decomposition of width at most \(w\), and
the running time is linear in the size of the supplied decomposition for fixed
\(w\).
\end{theorem}

\section{Bag-changing paths}

\begin{lemma}[Changing one bag into another]\label{lem:bag-path}
Let \(A,B\subseteq V(G)\) be finite sets with
\(|A|,|B|\le \omega+1\).  There is a rooted path \(P(A,B)\) whose top bag is
\(A\), whose bottom bag is \(B\), and whose consecutive bags differ by exactly
one vertex, except that if \(A=B\) the path consists of a single node.  Every
bag on the path has size at most \(\omega+1\), and the number of edges of the
path is
\[
  |A\setminus B|+|B\setminus A|\le 2(\omega+1).
\]
Moreover, for every vertex \(v\), the nodes of \(P(A,B)\) whose bags contain
\(v\) form a connected subpath, possibly empty.
\end{lemma}

\begin{proof}
Choose enumerations
\[
  A\setminus B=\{a_1,\ldots,a_p\},
  \qquad
  B\setminus A=\{b_1,\ldots,b_q\}.
\]
Start with \(A\).  First delete the vertices \(a_1,\ldots,a_p\), one at a time,
obtaining
\[
  A,\quad
  A\setminus\{a_1\},\quad
  A\setminus\{a_1,a_2\},\quad \ldots,\quad A\cap B.
\]
Then add the vertices \(b_1,\ldots,b_q\), one at a time, obtaining
\[
  A\cap B,\quad
  (A\cap B)\cup\{b_1\},\quad \ldots,\quad B.
\]
This gives a path from \(A\) to \(B\).  Each step changes exactly one vertex,
unless \(p=q=0\), in which case the path has only the single bag \(A=B\).
During the deletion part every bag is a subset of \(A\), and during the
insertion part every bag is a subset of \(B\), so all bags have size at most
\(\omega+1\).  The number of edges is \(p+q\).

For a fixed vertex \(v\), there are four cases.  If \(v\in A\cap B\), then
\(v\) is never deleted and is present throughout the path.  If
\(v\in A\setminus B\), then \(v\) is present until its deletion step and absent
afterwards.  If \(v\in B\setminus A\), then \(v\) is absent until its insertion
step and present afterwards.  If \(v\notin A\cup B\), it is absent throughout.
In every case the occurrence set is a connected subpath or is empty.
\end{proof}

\section{Algorithm}

We now give the algorithm promised in Theorem~\ref{thm:nice-normalization}.
The algorithm assumes that the input is already a tree-decomposition; it does
not spend time checking the decomposition axioms.

\paragraph{Subroutine: path insertion.}
For finite bags \(A,B\), define \(\operatorname{Path}(A,B)\) as in
Lemma~\ref{lem:bag-path}: first delete the elements of \(A\setminus B\), one at
a time, and then insert the elements of \(B\setminus A\), one at a time.  The
implementation uses the following version.

\begin{quote}
\textbf{AttachPath}\((A,B,z)\): here \(z\) is either absent, or is an already
constructed node whose bag is \(B\).  Compute the top-to-bottom bag sequence
\[
  A=B_0,B_1,\ldots,B_\ell=B
\]
of \(\operatorname{Path}(A,B)\).  If \(z\) is present, create fresh nodes with
bags \(B_0,\ldots,B_{\ell-1}\), link them into a path, and make the last fresh
node point to \(z\).  If \(\ell=0\), return \(z\).  If \(z\) is absent, create
fresh nodes with bags \(B_0,\ldots,B_\ell\) and link them into a path.  Return
the top node.
\end{quote}

In other words, \(\textbf{AttachPath}(A,B,z)\) creates a path whose top bag is
\(A\) and whose bottom bag is \(B\), reusing \(z\) as the bottom node when
possible.

\paragraph{Full algorithm.}
\begin{enumerate}[label=\textbf{Step \arabic*.},leftmargin=*]
  \item Choose an arbitrary node \(r\in\Nodes(T)\).  Root \(T\) at \(r\), compute
    the parent of each node, and list the nodes in postorder.  For a node \(t\),
    write \(\children(t)\) for its children in this rooted tree.

  \item Process the original nodes in postorder.  For each original node \(t\),
    construct a rooted output tree \(D_t\), and store its root node
    \(\rho(t)\).  The intended invariant is that \(\rho(t)\) has bag
    \(\beta(t)\), every leaf of \(D_t\) has empty bag, and \(D_t\) represents
    the part of the original decomposition below \(t\).

  \item If \(t\) has no children, set
    \[
      \rho(t) \defeq \textbf{AttachPath}(\beta(t),\emptyset,\text{absent}).
    \]

  \item If \(t\) has exactly one child \(s\), set
    \[
      \rho(t) \defeq \textbf{AttachPath}(\beta(t),\beta(s),\rho(s)).
    \]
    When \(\beta(t)=\beta(s)\), this creates no new same-bag unary node and
    simply returns \(\rho(s)\).

  \item If \(t\) has children \(s_1,\ldots,s_d\) with \(d\ge 2\), first form
    branches
    \[
      b_i \defeq \textbf{AttachPath}(\beta(t),\beta(s_i),\rho(s_i))
      \qquad (1\le i\le d).
    \]
    Each \(b_i\) has bag \(\beta(t)\).  Combine these branches by a
    right-associated binary join tree.  Concretely, start with \(J=b_d\).  For
    \(i=d-1,d-2,\ldots,1\), create a fresh node \(j_i\) with bag \(\beta(t)\),
    make \(b_i\) its first child and \(J\) its second child, and then reset
    \(J:=j_i\).  At the end, set \(\rho(t):=J\).

  \item After all original nodes have been processed, attach an empty-root path
    above \(\rho(r)\):
    \[
      r^\ast \defeq \textbf{AttachPath}(\emptyset,\beta(r),\rho(r)).
    \]
    Output the rooted tree with root \(r^\ast\), the child pointers just
    constructed, and the bag attached to every output node.
\end{enumerate}

The mathematical construction used in the verification below is exactly this
algorithm.

\section{Verification}

\begin{lemma}\label{lem:nice-local}
The pair \((T^\ast,\beta^\ast)\) is nice in the local sense of
Definition~1.1: the root and all leaves have empty bags, every node has at most
two children, unary nodes change one vertex, and binary nodes are join nodes.
All bags have size at most \(\omega+1\).
\end{lemma}

\begin{proof}
The new root is the top of \(P(\emptyset,\beta(r))\), so its bag is empty.
Leaves occur only as the bottom nodes of paths \(P(\beta(t),\emptyset)\) for
original leaves \(t\), and hence have empty bags.

Every path node has at most one child, and by Lemma~\ref{lem:bag-path} adjacent
bags on such a path differ by exactly one vertex.  The only nodes with two
children are the internal nodes of the binary join trees, and by construction
they and their two children all have bag \(\beta(t)\) for the relevant original
node \(t\).  The same lemma gives the width bound for all path nodes; join
nodes have bags equal to original bags, so they also have size at most
\(\omega+1\).
\end{proof}

\begin{lemma}\label{lem:original-bags-survive}
For every original node \(t\in\Nodes(T)\), some node
\(x\in\Nodes(T^\ast)\) satisfies \(\beta^\ast(x)=\beta(t)\).
\end{lemma}

\begin{proof}
We prove the corresponding statement inside each \(D_t\).  If \(t\) is a leaf,
the top node of \(P(\beta(t),\emptyset)\) has bag \(\beta(t)\).  If \(t\) has
one child \(s\) and \(\beta(t)\ne\beta(s)\), the top node of
\(P(\beta(t),\beta(s))\) has bag \(\beta(t)\).  If \(t\) has one child \(s\)
and \(\beta(t)=\beta(s)\), then the induction hypothesis for \(s\) gives a
node with bag \(\beta(s)=\beta(t)\).  If \(t\) has at least two children, the
root of the binary join tree has bag \(\beta(t)\).
\end{proof}

\begin{lemma}\label{lem:decomp-axioms}
The pair \((T^\ast,\beta^\ast)\) is a tree-decomposition of \(G\).
\end{lemma}

\begin{proof}
The vertex-covering axiom follows from Lemma~\ref{lem:original-bags-survive}
and the vertex-covering axiom for \((T,\beta)\).  Indeed, if \(v\in V(G)\),
choose \(t\in\Nodes(T)\) with \(v\in\beta(t)\); then a node of \(T^\ast\) has
bag \(\beta(t)\), hence contains \(v\).  The edge-covering axiom is identical:
if \(uv\in E(G)\), choose \(t\) with \(u,v\in\beta(t)\), and use the surviving
copy of \(\beta(t)\).

It remains to prove the running-intersection axiom.  Fix \(v\in V(G)\).  Let
\[
  S_v\defeq\{t\in\Nodes(T):v\in\beta(t)\}.
\]
This set is connected in the original tree \(T\).  We compare the occurrence
set of \(v\) in \(T^\ast\) with \(S_v\).

Each added path \(P(\beta(t),\beta(s))\) above a child \(s\) contributes a
connected subpath of nodes containing \(v\), by Lemma~\ref{lem:bag-path}.  If
\(v\) belongs to both endpoint bags, this subpath connects the side attached to
\(t\) with the side attached to \(s\).  If \(v\) belongs to exactly one endpoint
bag, the subpath is attached only at that endpoint.  If \(v\) belongs to
neither endpoint bag, the path contributes nothing.  The same description holds
for the top path \(P(\emptyset,\beta(r))\) and for the leaf paths
\(P(\beta(t),\emptyset)\).

Every join tree introduced at an original node \(t\) has all bags equal to
\(\beta(t)\).  Hence it is entirely contained in the occurrence set of \(v\)
when \(t\in S_v\), and it is disjoint from that occurrence set when
\(t\notin S_v\).

Thus the occurrence set of \(v\) in \(T^\ast\) is obtained from the connected
original subtree \(S_v\) by replacing original nodes and original edges with
connected pieces, and by attaching additional connected initial or terminal
path segments at nodes of \(S_v\).  Such replacements and attachments preserve
connectedness.  Therefore
\[
  \{x\in\Nodes(T^\ast):v\in\beta^\ast(x)\}
\]
is connected.
\end{proof}

\begin{proof}[Proof of Theorem~\ref{thm:nice-normalization}]
By Lemmas~\ref{lem:nice-local} and~\ref{lem:decomp-axioms},
\((T^\ast,\beta^\ast)\) is a nice tree-decomposition of \(G\) with all bags of
size at most \(\omega+1\).  It remains to record the size bound and the running
time.

Let \(n=|\Nodes(T)|\).  Each original edge \(ts\), oriented from parent \(t\)
to child \(s\), contributes at most \(2(\omega+1)\) new path nodes above
\(D_s\), because the bottom node is identified with the root of \(D_s\).
There are \(n-1\) original edges.  Each original leaf \(t\) contributes at most
\(\omega+2\) nodes in the terminal path \(P(\beta(t),\emptyset)\).  There are
at most \(n\) original leaves.  The binary join trees contribute
\(\max(d(t)-1,0)\) join nodes at a node with \(d(t)\) children, and
\[
  \sum_{t\in\Nodes(T)}\max(d(t)-1,0)
  \le \sum_{t\in\Nodes(T)} d(t)
  =n-1.
\]
Finally, the top path \(P(\emptyset,\beta(r))\) contributes at most
\(\omega+1\) new nodes above \(D_r\).

Therefore
\[
\begin{aligned}
  |\Nodes(T^\ast)|
  &\le 2(\omega+1)(n-1)+(\omega+2)n+(n-1)+(\omega+1) \\
  &= (3\omega+5)n-\omega-2 \\
  &\le (3\omega+5)n.
\end{aligned}
\]
This proves the claimed size bound.

For the running time, root \(T\) and compute the postorder in \(O(n)\) time.
For every oriented original edge \(ts\), the algorithm computes
\(\beta(t)\setminus\beta(s)\) and \(\beta(s)\setminus\beta(t)\).  Since both
bags have size at most \(\omega+1\), this takes \(O(\omega)\) time using a
temporary marking table for one bag, or \(O(\omega^2)\) time by direct list
comparison.  The leaf paths and the top path are handled in the same way.
The number of join nodes created at \(t\) is at most \(d(t)-1\), and the total
number of such nodes is at most \(n-1\), so join construction is linear.

Thus the node-generation work is \(O(\omega n)\) with linear-time finite-set
operations on bags.  If the final output writes every bag as an explicit list,
then each output bag has length at most \(\omega+1\), and the output-writing
cost is \(O(\omega|\Nodes(T^\ast)|)=O(\omega^2 n)\).  For fixed \(\omega\), this
is \(O(n)\), as claimed.
\end{proof}

\begin{remark}
The constant \(3\omega+5\) is not intended to be optimal.  The important point
for Courcelle's theorem is that the width is not increased, and for fixed
\(\omega\) both the number of nodes and the construction time are linear in the
size of the supplied decomposition.
\end{remark}

\end{document}
