\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{enumitem}
\usepackage{xcolor}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem*{theorem*}{Theorem}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\TW}{Thatcher--Wright}
\newcommand{\Nat}{\mathbb N}
\newcommand{\powfin}{\mathcal P_{\mathrm{fin}}}
\newcommand{\SigmaSym}{\Sigma}
\newcommand{\rank}{\sigma}
\newcommand{\Term}[1]{T_{#1}}
\newcommand{\Aut}{\mathcal A}
\newcommand{\NAut}{\mathcal R}
\newcommand{\bh}{\mathrm{bh}}
\newcommand{\MSO}{\mathrm{MSO}}
\newcommand{\child}{\mathrm{child}}
\newcommand{\dom}{\mathrm{dom}}
\newcommand{\arity}{\mathrm{arity}}
\newcommand{\enc}{\mathrm{enc}}
\newcommand{\TODO}[1]{\textcolor{red}{\textbf{TODO:} #1}}

\title{Working TeX Extraction of Thatcher--Wright}
\author{For the Courcelle formalization project}
\date{}

\begin{document}
\maketitle

\begin{abstract}
This note extracts the part of Thatcher and Wright~\cite{TW} that is useful for
formalising the tree-automata ingredient in Courcelle's theorem.  The source PDF
is a scan with a noisy text layer, so the material below is a cleaned working
transcription/paraphrase rather than a line-by-line archival copy.  The goal is
to identify the definitions, lemmas, automaton constructions, and proof
obligations that should eventually become Lean objects.
\end{abstract}

\tableofcontents

\section{How This Fits Courcelle}

The proof of Courcelle's theorem needs the following finite-tree automata
principle.  This is the statement called Theorem~26 in
\texttt{lecture\_note\_expanded.tex}.

\begin{theorem}[Lecture-note Theorem 26: effective \(\MSO\)--tree-automaton
correspondence]\label{thm:mso-tree-automata}\label{thm:lecture-note-26}
Let \(\Sigma\) be a finite alphabet.  From an \(\MSO\) sentence \(\Phi\) over
\[
  \tau_\Sigma=\{\child_1,\child_2\}\cup\{P_a:a\in\Sigma\}
\]
one can compute a deterministic bottom-up finite tree automaton
\(\Aut_\Phi\) such that, for every \(\Sigma\)-tree, \(\Aut_\Phi\) accepts it iff
it satisfies \(\Phi\).  Running \(\Aut_\Phi\) takes time linear in the number of
nodes.
\end{theorem}

\TW{} prove the needed ingredients in a slightly different language.
Their ``generalized finite automata'' are finite algebras over a ranked
signature; their inputs are finite terms.  This is exactly the bottom-up
ranked-tree automata view.  Sections~2--3 give finite automata theory for terms:
determinisation, Boolean closure, projection, inverse projection, replacement,
emptiness, and regularity.  Section~4 performs the logic-to-automata induction
for weak monadic second-order logic over the binary successor tree.

For our Courcelle notes, the encoding of finite subsets of \(N_2\) in
\S\,4 can mostly be replaced by labelled input trees directly.  The core proof
pattern is still the same: atomic formulas are recognizable, and recognizable
relations are closed under conjunction, negation, and existential set
quantification.

To connect the displayed theorem with \TW{}'s ranked-term formalism, encode a
\(\Sigma\)-tree as a full binary ranked term by adding a fresh nullary symbol
\(\bot\).  Each real node labelled \(a\in\Sigma\) becomes a binary symbol
\(a(t_1,t_2)\); an absent child is represented by \(\bot\).  The original
relations \(\child_1,\child_2\) and predicates \(P_a\) are first-order definable
on the non-\(\bot\) part of this padded term.  The \(\MSO\)-to-recognizable
induction below therefore produces a bottom-up automaton for the padded ranked
terms.  Determinization gives the deterministic automaton in
Theorem~\ref{thm:mso-tree-automata}, and its run is linear because it performs
one transition per node of a padded term whose size is \(O(|T|)\).

\medskip
\noindent\textbf{Derivation of Theorem~\ref{thm:mso-tree-automata} from \TW{}.}
Let \(\Gamma_\Sigma\) be the ranked alphabet with one nullary symbol \(\bot\)
and one binary symbol \(\widehat a\) for each \(a\in\Sigma\).  A finite
\(\Sigma\)-labelled binary tree \(T\) is represented by the
\(\Gamma_\Sigma\)-term \(\enc(T)\) obtained by putting \(\widehat a\) at every
real \(a\)-labelled node and \(\bot\) at every missing child.

Now let
\[
  n=1+|\Sigma|.
\]
The first track will represent the domain \(D\) of real nodes, and the remaining
tracks will represent the label sets \(P_a\).  Let \(\Sigma_n\) be the
\(n\)-track alphabet from Section~8.  Define an arity-preserving map
\[
  \iota:\Gamma_\Sigma\to\Sigma_n
\]
by
\[
  \iota(\bot)=\lambda,
\]
and
\[
  \iota(\widehat a)=\ell_a,
\]
where \(\ell_a\in\{0,1\}^n\) has domain-track value \(1\), has \(a\)-track value
\(1\), and has every other label-track value \(0\).  Then
\[
  \iota(\enc(T))=e(D_T,(P_a^T)_{a\in\Sigma}),
\]
where \(D_T\subseteq N_2\) is the finite prefix-closed set of positions of
nodes of \(T\), and \(P_a^T\subseteq D_T\) is the set of positions labelled
\(a\).  This encoding is canonical because absent subtrees are represented by
\(\lambda\), and every real node has domain-track value \(1\).

Given a tree \(\MSO\) sentence \(\Phi\) over
\(\{\child_1,\child_2\}\cup\{P_a:a\in\Sigma\}\), form a weak MSO formula
\[
  \Theta_\Phi(D,(P_a)_{a\in\Sigma})
\]
over \(N_2\) as follows.  Add a well-formedness formula saying that \(D\) is a
finite prefix-closed binary-tree domain and that the \(P_a\)'s partition \(D\).
Then relativize \(\Phi\) to \(D\): first-order quantifiers range over elements of
\(D\), set quantifiers range over subsets of \(D\), the child predicates are the
successor predicates \(R_i\) of \(N_2\), and each label predicate \(P_a(x)\) is
interpreted as \(x\in P_a\).  By induction on \(\Phi\),
\[
  T\models\Phi
  \quad\Longleftrightarrow\quad
  (D_T,(P_a^T)_{a\in\Sigma})\models\Theta_\Phi.        \tag{1}
\]

By the effective MSO-to-automata result over \(N_2\), namely
Theorem~\ref{thm:tw-definable-recognizable} together with the first-order
elimination of Section~\ref{sec:tw-eliminate-first-order}, the set
\[
  L_\Phi=
  \{e(D,(P_a)_{a\in\Sigma}):
      (D,(P_a)_{a\in\Sigma})\models\Theta_\Phi\}
  \subseteq\Term{\Sigma_n}
\]
is recognizable, effectively in \(\Phi\).  By closure under inverse projection
(Theorem~\ref{thm:tw-inverse-projection}),
\[
  \iota^{-1}(L_\Phi)\subseteq\Term{\Gamma_\Sigma}
\]
is recognizable.  By (1), this language contains exactly the padded terms
\(\enc(T)\) for finite \(\Sigma\)-trees satisfying \(\Phi\).  Finally,
Theorem~\ref{thm:tw-subset} determinizes the resulting automaton.  The
deterministic bottom-up run performs one transition per symbol of \(\enc(T)\),
and \(|\enc(T)|\le 2|T|+1\), so the running time is linear in the number of
nodes of \(T\).

\section{Generalized Automata}

\begin{definition}[Species and terms, TW \S2]
A \emph{species} is a pair \((\SigmaSym,\rank)\), where \(\SigmaSym\) is a set of
function symbols and \(\rank:\SigmaSym\to\Nat\) assigns each symbol its arity.
Write
\[
  \SigmaSym_n=\rank^{-1}(\{n\}).
\]
The set \(\Term{\SigmaSym}\) of finite terms is the least set such that
\(\SigmaSym_0\subseteq\Term{\SigmaSym}\) and, if \(f\in\SigmaSym_n\) and
\(t_1,\ldots,t_n\in\Term{\SigmaSym}\), then \(f(t_1,\ldots,t_n)\in
\Term{\SigmaSym}\).
\end{definition}

\begin{definition}[Deterministic automata]
A deterministic \(\SigmaSym\)-automaton is a finite
\(\SigmaSym\)-algebra
\[
  \Aut=(A,\alpha),
\]
where \(A\) is a finite set of states and, for each \(f\in\SigmaSym_n\),
\(\alpha_f:A^n\to A\).  The constants \(f\in\SigmaSym_0\) are initial states.
The bottom-up run is the unique homomorphism
\[
  h_\Aut:\Term{\SigmaSym}\to A
\]
defined by
\[
  h_\Aut(f(t_1,\ldots,t_n))
  =\alpha_f(h_\Aut(t_1),\ldots,h_\Aut(t_n)).
\]
For final states \(A_F\subseteq A\), the behaviour is
\[
  \bh_\Aut(A_F)=\{t\in\Term{\SigmaSym}:h_\Aut(t)\in A_F\}.
\]
\end{definition}

\begin{definition}[Nondeterministic automata]
A nondeterministic \(\SigmaSym\)-automaton is a finite relational algebra
\[
  \NAut=(R,\rho),
\]
where, for \(f\in\SigmaSym_n\), \(\rho_f\subseteq R^n\times R\).  Its run returns
a set of possible states:
\[
  h_\NAut(f(t_1,\ldots,t_n))
  =
  \{r:\exists r_1,\ldots,r_n,\ r_i\in h_\NAut(t_i)
      \text{ and } \rho_f(r_1,\ldots,r_n,r)\}.
\]
The behaviour with final states \(R_F\subseteq R\) is
\[
  \bh_\NAut(R_F)=\{t:h_\NAut(t)\cap R_F\neq\emptyset\}.
\]
\end{definition}

\begin{definition}[Recognizable sets]
A set \(U\subseteq\Term{\SigmaSym}\) is \emph{recognizable} if
\(U=\bh_\Aut(A_F)\) for some finite deterministic or nondeterministic
\(\SigmaSym\)-automaton.
\end{definition}

\begin{theorem}[TW Theorem 1: subset construction]\label{thm:tw-subset}
A set of terms is recognizable by a deterministic automaton iff it is
recognizable by a nondeterministic automaton.
\end{theorem}

\begin{proof}[Construction]
Every deterministic automaton is nondeterministic by replacing each transition
function by its graph.  More explicitly, if \(\Aut=(A,\alpha)\) is deterministic,
define a nondeterministic automaton on the same state set by
\[
  \rho_f(a_1,\ldots,a_n,a)
  \quad\Longleftrightarrow\quad
  \alpha_f(a_1,\ldots,a_n)=a
\]
for every \(f\in\SigmaSym_n\).  For \(n=0\) this says that the only possible
state at the constant \(f\) is \(\alpha_f\).  A structural induction on \(t\)
shows that the nondeterministic run is always the singleton
\[
  h_\NAut(t)=\{h_\Aut(t)\},
\]
so the same final states recognize the same language.

Conversely, let \(\NAut=(R,\rho)\) be a nondeterministic automaton.  Define the
deterministic powerset automaton
\[
  \widehat{\NAut}=(\mathcal P(R),\widehat\rho)
\]
as follows.  For \(f\in\SigmaSym_n\), set
\[
  \widehat\rho_f(S_1,\ldots,S_n)
  =
  \{r\in R:\exists r_1\in S_1,\ldots,\exists r_n\in S_n,\,
      \rho_f(r_1,\ldots,r_n,r)\}.
\]
When \(n=0\), the right hand side is read using the unique empty tuple:
\[
  \widehat\rho_f
  =
  \{r\in R:\rho_f(r)\}.
\]
Since \(R\) is finite, \(\mathcal P(R)\) is finite, so this is again a finite
automaton.  If \(R_F\) is the nondeterministic accepting set, take the accepting
states of the powerset automaton to be
\[
  \widehat R_F=\{S\subseteq R:S\cap R_F\neq\emptyset\}.
\]

It remains to prove that this construction has the same run semantics as the
original nondeterministic automaton.  We prove, by induction on terms, that
\[
  h_{\widehat{\NAut}}(t)=h_\NAut(t)       \tag{\(*\)}
\]
for every \(t\in\Term{\SigmaSym}\).

For the base case, let \(t=f\) with \(f\in\SigmaSym_0\).  By the definition of
the deterministic run in \(\widehat{\NAut}\),
\[
  h_{\widehat{\NAut}}(f)=\widehat\rho_f
  =
  \{r\in R:\rho_f(r)\}.
\]
This is exactly the definition of the nondeterministic run \(h_\NAut(f)\) at the
constant \(f\), so \((*)\) holds.

For the inductive step, let \(t=f(t_1,\ldots,t_n)\), and assume
\[
  h_{\widehat{\NAut}}(t_i)=h_\NAut(t_i)
  \qquad (1\le i\le n).
\]
Then
\begin{align*}
  h_{\widehat{\NAut}}(t)
  &=
  \widehat\rho_f\bigl(h_{\widehat{\NAut}}(t_1),\ldots,
    h_{\widehat{\NAut}}(t_n)\bigr) \\
  &=
  \{r\in R:\exists r_1\in h_{\widehat{\NAut}}(t_1),\ldots,
      \exists r_n\in h_{\widehat{\NAut}}(t_n),\,
      \rho_f(r_1,\ldots,r_n,r)\} \\
  &=
  \{r\in R:\exists r_1\in h_\NAut(t_1),\ldots,
      \exists r_n\in h_\NAut(t_n),\,
      \rho_f(r_1,\ldots,r_n,r)\} \\
  &=
  h_\NAut(f(t_1,\ldots,t_n)).
\end{align*}
The third equality is the induction hypothesis, and the last equality is the
definition of \(h_\NAut\).  This proves \((*)\) for all terms.

Finally, for every term \(t\),
\[
  t\in\bh_{\widehat{\NAut}}(\widehat R_F)
  \Longleftrightarrow
  h_{\widehat{\NAut}}(t)\cap R_F\neq\emptyset
  \Longleftrightarrow
  h_\NAut(t)\cap R_F\neq\emptyset
  \Longleftrightarrow
  t\in\bh_\NAut(R_F).
\]
Thus the powerset automaton recognizes exactly the same set of terms as
\(\NAut\), completing the nontrivial direction.
\end{proof}

\begin{theorem}[TW Theorem 2: Boolean closure]\label{thm:tw-boolean}
If \(U,V\subseteq\Term{\SigmaSym}\) are recognizable, then so are
\(U\cap V\), \(U\cup V\), and \(\Term{\SigmaSym}\setminus U\).
\end{theorem}

\begin{proof}[Construction]
By Theorem~\ref{thm:tw-subset}, we may assume the recognizing automata are
deterministic.  Let
\[
  \Aut=(A,\alpha),\qquad \mathcal B=(B,\beta)
\]
be deterministic \(\SigmaSym\)-automata with final-state sets
\(A_F\subseteq A\) and \(B_F\subseteq B\), and suppose
\[
  \bh_\Aut(A_F)=U,\qquad \bh_{\mathcal B}(B_F)=V.
\]

\emph{Complement.}  Use the same automaton \(\Aut\), but replace the accepting
set \(A_F\) by \(A\setminus A_F\).  For every term \(t\),
\[
  t\in\bh_\Aut(A\setminus A_F)
  \Longleftrightarrow
  h_\Aut(t)\in A\setminus A_F
  \Longleftrightarrow
  h_\Aut(t)\notin A_F
  \Longleftrightarrow
  t\notin\bh_\Aut(A_F).
\]
Thus
\[
  \bh_\Aut(A\setminus A_F)=\Term{\SigmaSym}\setminus U.
\]
This is why determinism is useful here: each term has one state, so changing the
final set really complements the language.  For a nondeterministic automaton,
one must first determinize.

\emph{Intersection.}  Define the product automaton
\[
  \mathcal C=(A\times B,\gamma).
\]
For a constant \(f\in\SigmaSym_0\), put
\[
  \gamma_f=(\alpha_f,\beta_f)\in A\times B.
\]
For \(f\in\SigmaSym_n\) with \(n>0\), define
\[
  \gamma_f\bigl((a_1,b_1),\ldots,(a_n,b_n)\bigr)
  =
  \bigl(\alpha_f(a_1,\ldots,a_n),\beta_f(b_1,\ldots,b_n)\bigr).
\]
Equivalently, the same formula also covers \(n=0\) if one reads the empty tuple
correctly.

The crucial fact is that the product automaton computes both original runs at
the same time:
\[
  h_{\mathcal C}(t)=\bigl(h_\Aut(t),h_{\mathcal B}(t)\bigr)       \tag{\(*\)}
\]
for every \(t\in\Term{\SigmaSym}\).  We prove this by structural induction on
\(t\).

If \(t=f\) is a constant, then
\[
  h_{\mathcal C}(f)=\gamma_f=(\alpha_f,\beta_f)
  =
  \bigl(h_\Aut(f),h_{\mathcal B}(f)\bigr).
\]
For the inductive step, let \(t=f(t_1,\ldots,t_n)\), and assume \((*)\) holds
for each \(t_i\).  Then
\begin{align*}
  h_{\mathcal C}(t)
  &=
  \gamma_f\bigl(h_{\mathcal C}(t_1),\ldots,h_{\mathcal C}(t_n)\bigr) \\
  &=
  \gamma_f\bigl((h_\Aut(t_1),h_{\mathcal B}(t_1)),\ldots,
    (h_\Aut(t_n),h_{\mathcal B}(t_n))\bigr) \\
  &=
  \bigl(\alpha_f(h_\Aut(t_1),\ldots,h_\Aut(t_n)),
        \beta_f(h_{\mathcal B}(t_1),\ldots,h_{\mathcal B}(t_n))\bigr) \\
  &=
  \bigl(h_\Aut(f(t_1,\ldots,t_n)),
        h_{\mathcal B}(f(t_1,\ldots,t_n))\bigr).
\end{align*}
This proves \((*)\).

Now take final states \(A_F\times B_F\) in the product automaton.  Using
\((*)\), for every term \(t\),
\begin{align*}
  t\in\bh_{\mathcal C}(A_F\times B_F)
  &\Longleftrightarrow
  h_{\mathcal C}(t)\in A_F\times B_F \\
  &\Longleftrightarrow
  h_\Aut(t)\in A_F \text{ and } h_{\mathcal B}(t)\in B_F \\
  &\Longleftrightarrow
  t\in U \text{ and } t\in V.
\end{align*}
Therefore
\[
  \bh_{\mathcal C}(A_F\times B_F)=U\cap V.
\]
This proves that the product automaton with accepting set \(A_F\times B_F\) is
the desired automaton for intersection.

\emph{Union.}  Use the same product automaton \(\mathcal C\), but take accepting
states
\[
  (A_F\times B)\cup(A\times B_F).
\]
Again using \((*)\), for every term \(t\),
\begin{align*}
  t\in\bh_{\mathcal C}\bigl((A_F\times B)\cup(A\times B_F)\bigr)
  &\Longleftrightarrow
  h_\Aut(t)\in A_F \text{ or } h_{\mathcal B}(t)\in B_F \\
  &\Longleftrightarrow
  t\in U \text{ or } t\in V.
\end{align*}
Hence this product automaton recognizes \(U\cup V\).  This also follows from
complement and intersection by De Morgan's laws, but the direct final-state set
makes the automaton construction explicit.
\end{proof}

\section{Projection}

\begin{definition}[Projection of species]
Let \((\SigmaSym,\rank)\) and \((\Gamma,\rank')\) be species.  A map
\(\pi:\SigmaSym\to\Gamma\) is a projection if it preserves arities:
\[
  \rank'(\pi(f))=\rank(f).
\]
It extends to terms by
\[
  \pi(f(t_1,\ldots,t_n))
  =
  \pi(f)(\pi(t_1),\ldots,\pi(t_n)).
\]
\end{definition}

\begin{theorem}[TW Theorem 3: closure under projection]\label{thm:tw-projection}
If \(U\subseteq\Term{\SigmaSym}\) is recognizable and
\(\pi:\Term{\SigmaSym}\to\Term{\Gamma}\) is induced by an arity-preserving symbol
map, then \(\pi(U)\subseteq\Term{\Gamma}\) is recognizable.
\end{theorem}

\begin{proof}[Construction]
By Theorem~\ref{thm:tw-subset}, assume \(U\) is recognized by a deterministic
\(\SigmaSym\)-automaton
\[
  \Aut=(A,\alpha)
\]
with final states \(A_F\).  We build a nondeterministic \(\Gamma\)-automaton
\[
  \mathcal N=(A,\eta)
\]
on the same state set.  Its transitions are obtained by forgetting which
\(\SigmaSym\)-symbol was used before projection.

For a constant \(g\in\Gamma_0\), define the possible initial states by
\[
  \eta_g
  =
  \{\alpha_f:f\in\SigmaSym_0\text{ and }\pi(f)=g\}.
\]
For \(g\in\Gamma_n\) with \(n>0\), define
\[
  \eta_g(a_1,\ldots,a_n,a)
  \quad\Longleftrightarrow\quad
  \exists f\in\SigmaSym_n,\ \pi(f)=g
  \text{ and }
  \alpha_f(a_1,\ldots,a_n)=a.
\]
Thus a \(g\)-transition of \(\mathcal N\) is allowed exactly when it comes from
some source transition whose label projects to \(g\).  We use the same final
states \(A_F\) for \(\mathcal N\).

We prove two structural-induction facts.  The first says that every source run
survives after projection:
\[
  \text{for all }s\in\Term{\SigmaSym},\qquad
  h_\Aut(s)\in h_{\mathcal N}(\pi(s)).            \tag{P1}
\]
If \(s=f\) is a constant, then \(\pi(s)=\pi(f)\) and
\[
  h_\Aut(f)=\alpha_f\in\eta_{\pi(f)}=h_{\mathcal N}(\pi(f))
\]
by the definition of the initial states of \(\mathcal N\).  For the inductive
step, write \(s=f(s_1,\ldots,s_n)\), and assume
\[
  h_\Aut(s_i)\in h_{\mathcal N}(\pi(s_i))
  \qquad (1\le i\le n).
\]
Let
\[
  a_i=h_\Aut(s_i),\qquad
  a=h_\Aut(s)=\alpha_f(a_1,\ldots,a_n).
\]
Since \(\pi(f)\) is the label of the projected term and \(f\) is one of the
symbols projecting to \(\pi(f)\), the definition of \(\eta_{\pi(f)}\) gives
\[
  \eta_{\pi(f)}(a_1,\ldots,a_n,a).
\]
Together with the induction hypotheses \(a_i\in h_{\mathcal N}(\pi(s_i))\),
this says exactly that
\[
  a\in h_{\mathcal N}\bigl(\pi(f)(\pi(s_1),\ldots,\pi(s_n))\bigr)
  =
  h_{\mathcal N}(\pi(s)).
\]
So (P1) holds for \(s\).

The second fact is the converse lifting property: every state reached by the
projected nondeterministic automaton comes from some source term.
\[
  \text{for all }t\in\Term{\Gamma}\text{ and }a\in A,\qquad
  a\in h_{\mathcal N}(t)
  \Rightarrow
  \exists s\in\Term{\SigmaSym},\ \pi(s)=t\text{ and }h_\Aut(s)=a.       \tag{P2}
\]
Again prove this by induction on \(t\).  If \(t=g\) is a constant and
\(a\in h_{\mathcal N}(g)=\eta_g\), then by definition of \(\eta_g\) there is a
constant \(f\in\SigmaSym_0\) with \(\pi(f)=g\) and \(\alpha_f=a\).  Taking
\(s=f\), we get \(\pi(s)=g=t\) and \(h_\Aut(s)=a\).

For the inductive step, let \(t=g(t_1,\ldots,t_n)\), and suppose
\(a\in h_{\mathcal N}(t)\).  By the definition of the nondeterministic run,
there are states \(a_i\in h_{\mathcal N}(t_i)\) such that
\[
  \eta_g(a_1,\ldots,a_n,a).
\]
By the definition of \(\eta_g\), there is some \(f\in\SigmaSym_n\) with
\[
  \pi(f)=g
  \quad\text{and}\quad
  \alpha_f(a_1,\ldots,a_n)=a.
\]
By the induction hypothesis applied to \(t_i\) and \(a_i\), choose source terms
\(s_i\in\Term{\SigmaSym}\) such that
\[
  \pi(s_i)=t_i
  \quad\text{and}\quad
  h_\Aut(s_i)=a_i.
\]
Now define
\[
  s=f(s_1,\ldots,s_n).
\]
Then
\[
  \pi(s)
  =
  \pi(f)(\pi(s_1),\ldots,\pi(s_n))
  =
  g(t_1,\ldots,t_n)
  =
  t,
\]
and
\[
  h_\Aut(s)
  =
  \alpha_f(h_\Aut(s_1),\ldots,h_\Aut(s_n))
  =
  \alpha_f(a_1,\ldots,a_n)
  =
  a.
\]
This proves (P2).

We can now identify the language recognized by \(\mathcal N\).  First, if
\(t\in\pi(U)\), then \(t=\pi(s)\) for some \(s\in U\).  Since
\(s\in U=\bh_\Aut(A_F)\), we have \(h_\Aut(s)\in A_F\).  By (P1),
\[
  h_\Aut(s)\in h_{\mathcal N}(\pi(s))=h_{\mathcal N}(t),
\]
so \(h_{\mathcal N}(t)\cap A_F\neq\emptyset\), and \(t\in\bh_{\mathcal N}(A_F)\).

Conversely, if \(t\in\bh_{\mathcal N}(A_F)\), then choose
\[
  a\in h_{\mathcal N}(t)\cap A_F.
\]
By (P2), there is \(s\in\Term{\SigmaSym}\) with \(\pi(s)=t\) and
\(h_\Aut(s)=a\).  Since \(a\in A_F\), we have \(s\in U\), hence
\[
  t=\pi(s)\in\pi(U).
\]
Thus
\[
  \bh_{\mathcal N}(A_F)=\pi(U),
\]
so \(\pi(U)\) is recognizable.
\end{proof}

\begin{theorem}[TW Theorem 4: closure under inverse projection]\label{thm:tw-inverse-projection}
If \(U\subseteq\Term{\Gamma}\) is recognizable and
\(\pi:\Term{\SigmaSym}\to\Term{\Gamma}\) is induced by an arity-preserving symbol
map, then \(\pi^{-1}(U)\subseteq\Term{\SigmaSym}\) is recognizable.
\end{theorem}

\begin{proof}[Construction]
By Theorem~\ref{thm:tw-subset}, assume \(U\) is recognized by a deterministic
\(\Gamma\)-automaton
\[
  \Aut=(A,\alpha)
\]
with final states \(A_F\).  Define a deterministic \(\SigmaSym\)-automaton
\[
  \mathcal D=(A,\beta)
\]
on the same state set by reusing the transition of the projected symbol:
\[
  \beta_f=\alpha_{\pi(f)}
  \qquad (f\in\SigmaSym).
\]
This is well typed because \(\pi\) preserves arities.

The key invariant is
\[
  h_{\mathcal D}(s)=h_\Aut(\pi(s))
  \qquad\text{for every }s\in\Term{\SigmaSym}.       \tag{\(*\)}
\]
We prove it by structural induction on \(s\).  If \(s=f\) is a constant, then
\[
  h_{\mathcal D}(f)
  =
  \beta_f
  =
  \alpha_{\pi(f)}
  =
  h_\Aut(\pi(f)).
\]
For the inductive step, let \(s=f(s_1,\ldots,s_n)\), and assume
\[
  h_{\mathcal D}(s_i)=h_\Aut(\pi(s_i))
  \qquad (1\le i\le n).
\]
Then
\begin{align*}
  h_{\mathcal D}(s)
  &=
  \beta_f\bigl(h_{\mathcal D}(s_1),\ldots,h_{\mathcal D}(s_n)\bigr) \\
  &=
  \alpha_{\pi(f)}
    \bigl(h_\Aut(\pi(s_1)),\ldots,h_\Aut(\pi(s_n))\bigr) \\
  &=
  h_\Aut\bigl(\pi(f)(\pi(s_1),\ldots,\pi(s_n))\bigr) \\
  &=
  h_\Aut(\pi(s)).
\end{align*}
This proves \((*)\).

Use the same accepting set \(A_F\) for \(\mathcal D\).  For every
\(\SigmaSym\)-term \(s\),
\[
  s\in\bh_{\mathcal D}(A_F)
  \Longleftrightarrow
  h_{\mathcal D}(s)\in A_F
  \Longleftrightarrow
  h_\Aut(\pi(s))\in A_F
  \Longleftrightarrow
  \pi(s)\in U
  \Longleftrightarrow
  s\in\pi^{-1}(U).
\]
Therefore
\[
  \bh_{\mathcal D}(A_F)=\pi^{-1}(U),
\]
so the inverse image is recognizable.
\end{proof}

\section{Replacement and Emptiness}

\begin{definition}[Subterm and depth]
The depth of a term is
\[
  d(c)=1 \quad(c\in\SigmaSym_0), \qquad
  d(f(t_1,\ldots,t_n))=1+\max_i d(t_i).
\]
Write \(t\succeq s\) when \(s\) occurs as a subterm of \(t\).
\end{definition}

\begin{lemma}[TW Lemma 5: replacement]\label{lem:tw-replacement}
Let \(\Aut\) be a deterministic \(\SigmaSym\)-automaton.  Suppose \(t'\) is
obtained from \(t\) by replacing an occurrence of \(t_2\) by \(t_1\).  If
\[
  h_\Aut(t_1)=h_\Aut(t_2),
\]
then
\[
  h_\Aut(t')=h_\Aut(t).
\]
\end{lemma}

\begin{proof}
We prove the statement by structural induction on the term \(t\).  The induction
statement is: for every occurrence of \(t_2\) in \(t\), if \(t'\) is obtained by
replacing that occurrence by \(t_1\), and if \(h_\Aut(t_1)=h_\Aut(t_2)\), then
\(h_\Aut(t')=h_\Aut(t)\).

First suppose \(t=c\) is a constant symbol.  A constant term has only one
occurrence, namely the whole term itself.  Therefore the replaced occurrence must
be the root, so \(t_2=t=c\) and \(t'=t_1\).  Hence
\[
  h_\Aut(t')
  =
  h_\Aut(t_1)
  =
  h_\Aut(t_2)
  =
  h_\Aut(t).
\]

Now suppose
\[
  t=f(u_1,\ldots,u_n).
\]
There are two cases.

If the replaced occurrence is the root of \(t\), then \(t_2=t\) and \(t'=t_1\).
The same calculation as in the base case gives
\[
  h_\Aut(t')=h_\Aut(t_1)=h_\Aut(t_2)=h_\Aut(t).
\]

Otherwise the occurrence lies inside exactly one child \(u_j\).  Then there is a
term \(u'_j\) obtained from \(u_j\) by replacing that occurrence of \(t_2\) by
\(t_1\), and
\[
  t'=f(u_1,\ldots,u_{j-1},u'_j,u_{j+1},\ldots,u_n).
\]
By the induction hypothesis applied to \(u_j\),
\[
  h_\Aut(u'_j)=h_\Aut(u_j).
\]
Therefore
\begin{align*}
  h_\Aut(t')
  &=
  \alpha_f\bigl(h_\Aut(u_1),\ldots,h_\Aut(u_{j-1}),
    h_\Aut(u'_j),h_\Aut(u_{j+1}),\ldots,h_\Aut(u_n)\bigr) \\
  &=
  \alpha_f\bigl(h_\Aut(u_1),\ldots,h_\Aut(u_{j-1}),
    h_\Aut(u_j),h_\Aut(u_{j+1}),\ldots,h_\Aut(u_n)\bigr) \\
  &=
  h_\Aut(f(u_1,\ldots,u_n))
  =
  h_\Aut(t).
\end{align*}
This completes the induction.
\end{proof}

\begin{corollary}[TW corollary to Lemma 5]
Let \(u_1,\ldots,u_m\) be pairwise incomparable occurrences of subterms of a term
\(t\).  For each \(i\), let \(v_i\) be a term satisfying
\[
  h_\Aut(v_i)=h_\Aut(u_i).
\]
If \(t'\) is obtained from \(t\) by simultaneously replacing each occurrence
\(u_i\) by \(v_i\), then
\[
  h_\Aut(t')=h_\Aut(t).
\]
\end{corollary}

\begin{proof}
Enumerate the incomparable occurrences in any order.  Define a sequence of terms
\[
  t=t^{(0)},t^{(1)},\ldots,t^{(m)}=t',
\]
where \(t^{(k)}\) is obtained from \(t^{(k-1)}\) by replacing the \(k\)-th
selected occurrence \(u_k\) by \(v_k\).

The reason this sequential replacement is legitimate is exactly the
incomparability hypothesis.  Since no selected occurrence is inside another,
replacing one selected occurrence neither deletes nor changes any of the other
selected occurrences.  Thus, when we pass from \(t^{(k-1)}\) to \(t^{(k)}\), the
subterm being replaced is still the original \(u_k\), and its replacement \(v_k\)
has the same automaton state:
\[
  h_\Aut(v_k)=h_\Aut(u_k).
\]
Applying Lemma~\ref{lem:tw-replacement} to the \(k\)-th step gives
\[
  h_\Aut(t^{(k)})=h_\Aut(t^{(k-1)}).
\]
Chaining these equalities yields
\[
  h_\Aut(t')
  =
  h_\Aut(t^{(m)})
  =
  h_\Aut(t^{(m-1)})
  =
  \cdots
  =
  h_\Aut(t^{(0)})
  =
  h_\Aut(t).
\]
\end{proof}

\begin{lemma}[TW Lemma 6: bounded-depth witness]\label{lem:tw-bounded-depth}
Let \(\Aut\) have \(n\) states.  If some term \(t\) has \(h_\Aut(t)=a\), then
there is a term \(t'\) with \(d(t')\le n\) and \(h_\Aut(t')=a\).
\end{lemma}

\begin{proof}
Let \(|s|\) denote the number of symbols in a term \(s\).  Among all terms
reaching the state \(a\), choose one, say \(s\), of minimal size:
\[
  h_\Aut(s)=a
  \qquad\text{and}\qquad
  |s|\le |r| \text{ whenever } h_\Aut(r)=a.
\]
Such a term exists because the original term \(t\) reaches \(a\), and sizes are
natural numbers.

We claim that \(d(s)\le n\).  Suppose not, so \(d(s)>n\).  By the definition of
depth, there is a longest root-to-leaf chain of subterm occurrences
\[
  s=s_0 \succ s_1 \succ \cdots \succ s_{d(s)-1},
\]
where each \(s_{i+1}\) is a proper child-subterm somewhere below \(s_i\).  This
chain contains \(d(s)\) subterms.  Since \(d(s)>n\) and the automaton has only
\(n\) states, the pigeonhole principle gives indices \(i<j\) such that
\[
  h_\Aut(s_i)=h_\Aut(s_j).
\]
Here \(s_j\) is a proper subterm of \(s_i\), hence \(|s_j|<|s_i|\).

Let \(s^\star\) be the term obtained from \(s\) by replacing the occurrence
\(s_i\) by the smaller term \(s_j\).  Lemma~\ref{lem:tw-replacement}, applied
with \(t_1=s_j\) and \(t_2=s_i\), gives
\[
  h_\Aut(s^\star)=h_\Aut(s)=a.
\]
But the replacement strictly decreases size:
\[
  |s^\star|=|s|-|s_i|+|s_j|<|s|.
\]
This contradicts the minimality of \(s\) among terms reaching \(a\).  Therefore
\(d(s)\le n\).  Taking \(t'=s\) proves the lemma.
\end{proof}

\begin{theorem}[TW Theorem 7: decidable emptiness]\label{thm:tw-emptiness}
Given a finite \(\SigmaSym\)-automaton \(\Aut\) and final states \(A_F\), it is
decidable whether \(\bh_\Aut(A_F)=\emptyset\).
\end{theorem}

\begin{proof}
By Theorem~\ref{thm:tw-subset}, it is enough to describe the procedure for a
deterministic automaton.  Let \(n\) be the number of states of \(\Aut\).

First observe that the set of terms of depth at most \(n\) is finite and can be
effectively enumerated.  Define finite sets \(D_k\subseteq\Term{\SigmaSym}\) by
\[
  D_1=\SigmaSym_0,
\]
and
\[
  D_{k+1}
  =
  D_k\cup
  \{f(t_1,\ldots,t_m):
    f\in\SigmaSym_m,\ t_1,\ldots,t_m\in D_k\}.
\]
Because the species is finite, each \(\SigmaSym_m\) is finite, and because the
automaton has finite arities appearing in \(\SigmaSym\), each \(D_k\) is finite.
An induction on \(k\) shows that \(D_k\) is exactly the set of terms of depth at
most \(k\).

The decision procedure is now:
\begin{enumerate}[leftmargin=*]
  \item enumerate \(D_n\);
  \item compute \(h_\Aut(s)\) for every \(s\in D_n\);
  \item answer ``nonempty'' iff \(h_\Aut(s)\in A_F\) for some \(s\in D_n\).
\end{enumerate}

The procedure is sound: if it finds such an \(s\), then
\[
  s\in\bh_\Aut(A_F),
\]
so \(\bh_\Aut(A_F)\neq\emptyset\).

It is also complete.  Suppose \(\bh_\Aut(A_F)\neq\emptyset\).  Then there is a
term \(t\) such that \(h_\Aut(t)=a\in A_F\).  By
Lemma~\ref{lem:tw-bounded-depth}, there is a term \(t'\) with
\[
  d(t')\le n
  \qquad\text{and}\qquad
  h_\Aut(t')=a.
\]
Thus \(t'\in D_n\), and the finite search will find an accepting term.  Hence
the procedure answers ``empty'' exactly when \(\bh_\Aut(A_F)=\emptyset\).
\end{proof}

\section{Regularity}

\begin{definition}[Complex product and closure, TW \S3]
Fix an individual symbol \(\lambda\in\SigmaSym_0\).  For
\(U,V\subseteq\Term{\SigmaSym}\), the complex product
\[
  U\cdot_\lambda V
\]
is the set of all terms obtained from a term \(u\in U\) by replacing every
occurrence of \(\lambda\) by an independently chosen term of \(V\).

The \(\lambda\)-closure \(U^{*_\lambda}\) is defined by
\[
  X_0=\{\lambda\}, \qquad
  X_{n+1}=X_n\cup(U\cdot_\lambda X_n), \qquad
  U^{*_\lambda}=\bigcup_{n\ge0}X_n.
\]
\end{definition}

\begin{definition}[Regular sets of terms]
A set \(U\subseteq\Term{\SigmaSym}\) is regular if it belongs to the least class
containing all finite sets and closed under finite union, complex products, and
closures, after allowing the signature to be extended by additional constant
symbols but not by additional positive-arity symbols.
\end{definition}

\begin{theorem*}[TW Theorem 9: analysis theorem]
Every recognizable subset of \(\Term{\SigmaSym}\) is regular.
\end{theorem*}

\begin{proof}
Let \(U=\bh_\Aut(A_F)\), where \(\Aut=(A,\alpha)\) is a deterministic
\(\SigmaSym\)-automaton.  Write the finite state set as
\[
  A=\{a_1,\ldots,a_m\}.
\]
Introduce fresh constant symbols
\[
  \lambda_1,\ldots,\lambda_m,
\]
one for each state.  Let \(\Sigma^+\) be the species with the same
positive-arity symbols as \(\SigmaSym\), but with constants
\(\lambda_1,\ldots,\lambda_m\).  Define a \(\Sigma^+\)-automaton
\(\Aut^+=(A,\alpha^+)\) by
\[
  \alpha^+_{\lambda_i}=a_i,
  \qquad
  \alpha^+_f=\alpha_f
  \quad(f\in\SigmaSym_n,\ n>0).
\]
The fresh constant \(\lambda_i\) should be read as a placeholder for ``some
term whose automaton state is \(a_i\).''

For \(j=1,\ldots,m\), let
\[
  L_j=\{t\in\Term{\Sigma^+}:h_{\Aut^+}(t)=a_j\}.
\]
We first prove that every \(L_j\) is regular.  This is the analysis part of
TW's proof.

For each state \(a_j\), define the finite set of one-step transition patterns
\[
  U_j=
  \{f(\lambda_{i_1},\ldots,\lambda_{i_r}):
      f\in\SigmaSym_r,\ r>0,
      \alpha_f(a_{i_1},\ldots,a_{i_r})=a_j\}.
\]
Each \(U_j\) is finite, hence regular.

We now approximate \(\Aut^+\) by allowing non-leaf transitions only into the
first \(k\) states.  Define \(T_j^k\) to be the set of \(\Sigma^+\)-terms whose
\(\Aut^+\)-run has root state \(a_j\) and in which every non-leaf transition
result belongs to \(\{a_1,\ldots,a_k\}\).  Constants \(\lambda_i\) are always
allowed, since they are leaves rather than non-leaf transitions.  Thus
\[
  T_j^m=L_j.
\]
We prove by induction on \(k\) that all sets \(T_j^k\) are regular.

For \(k=1\), no non-leaf transition may produce any state except \(a_1\).  If
\(j\ne1\), the only way to get state \(a_j\) is the leaf \(\lambda_j\), so
\[
  T_j^1=\{\lambda_j\}.
\]
For \(j=1\), a term in \(T_1^1\) is obtained by iterating transition patterns
from \(U_1\) at holes labelled \(\lambda_1\).  In other words,
\[
  T_1^1=U_1^{*_{\lambda_1}}.
\]
The inclusion \(U_1^{*_{\lambda_1}}\subseteq T_1^1\) follows by induction on
the closure stage: every pattern in \(U_1\) has root state \(a_1\), and replacing
a \(\lambda_1\)-hole by a term whose state is \(a_1\) preserves the surrounding
state by the replacement lemma.  Conversely, induction on the term shows that
any \(T_1^1\)-term is built by exactly this iteration, since every non-leaf node
must be one of the patterns in \(U_1\).  Hence the base case is regular.

Assume now that all \(T_j^k\) are regular.  We construct the sets \(T_j^{k+1}\).
First form
\[
  V_{k+1}
  =
  (((U_{k+1}\cdot_{\lambda_1}T_1^k)
      \cdot_{\lambda_2}T_2^k)\cdots
      \cdot_{\lambda_m}T_m^k).
\]
The set \(V_{k+1}\) consists of terms whose root is one new non-leaf transition
to \(a_{k+1}\), while all proper subcomputations use only the first \(k\) states
as non-leaf transition results.  By the induction hypothesis and closure of
regular sets under complex product, \(V_{k+1}\) is regular.  Let
\[
  W_{k+1}=V_{k+1}^{*_{\lambda_{k+1}}}.
\]
This is regular by closure under \(\lambda_{k+1}\)-closure.

We claim that, for \(j\le k+1\),
\[
  T_j^{k+1}=T_j^k\cdot_{\lambda_{k+1}}W_{k+1}.       \tag{\(*\)}
\]
The inclusion from right to left is the easy replacement direction.  Every term
of \(W_{k+1}\) has \(\Aut^+\)-state \(a_{k+1}\), proved by induction on the
closure stage defining \(W_{k+1}\): each element of \(V_{k+1}\) has root state
\(a_{k+1}\), and replacing \(\lambda_{k+1}\)-holes by terms of state
\(a_{k+1}\) preserves state.  Therefore replacing \(\lambda_{k+1}\)-leaves in a
term from \(T_j^k\) by elements of \(W_{k+1}\) preserves the root state \(a_j\)
and introduces only allowed non-leaf transitions to \(a_{k+1}\).  Hence the
result lies in \(T_j^{k+1}\).

For the converse inclusion, let \(X_0=\{\lambda_{k+1}\}\) and
\[
  X_{r+1}=X_r\cup (V_{k+1}\cdot_{\lambda_{k+1}}X_r).
\]
Thus \(W_{k+1}=\bigcup_r X_r\).  We prove, by induction on the number \(r\) of
non-leaf transitions to \(a_{k+1}\) occurring in the computation of a term
\(t\in T_j^{k+1}\), that
\[
  t\in T_j^k\cdot_{\lambda_{k+1}}X_r.        \tag{\(**\)}
\]
If \(r=0\), then no non-leaf transition to \(a_{k+1}\) occurs, so the same
computation already satisfies the definition of \(T_j^k\).  Hence
\[
  t\in T_j^k=T_j^k\cdot_{\lambda_{k+1}}\{\lambda_{k+1}\}
  =T_j^k\cdot_{\lambda_{k+1}}X_0.
\]
For the inductive step, choose a lowest occurrence whose root transition is a
non-leaf transition to \(a_{k+1}\).  Call this subterm \(v\).  Its immediate
children have no non-leaf transition to \(a_{k+1}\) below them, so each child
belongs to some \(T_i^k\).  Therefore \(v\in V_{k+1}\).  Replace this occurrence
of \(v\) in \(t\) by \(\lambda_{k+1}\), obtaining a term \(t_0\).  The term
\(t_0\) has one fewer non-leaf transition to \(a_{k+1}\), so by induction
\[
  t_0\in T_j^k\cdot_{\lambda_{k+1}}X_r.
\]
Restoring \(v\) means replacing one \(\lambda_{k+1}\)-hole by an element of
\(V_{k+1}\), hence
\[
  t\in (T_j^k\cdot_{\lambda_{k+1}}X_r)
        \cdot_{\lambda_{k+1}}V_{k+1}
  \subseteq T_j^k\cdot_{\lambda_{k+1}}X_{r+1}.
\]
This proves \((**)\), and therefore \(T_j^{k+1}\subseteq
T_j^k\cdot_{\lambda_{k+1}}W_{k+1}\).  The claim \((*)\) follows.

For \(j>k+1\), still no non-leaf transition may produce \(a_j\), so
\[
  T_j^{k+1}=\{\lambda_j\}.
\]
Thus all \(T_j^{k+1}\) are regular, completing the induction on \(k\).  Hence
all \(L_j=T_j^m\) are regular.

It remains to translate this back from the fresh constants to the original
constants of \(\SigmaSym\).  For each state \(a_i\), let
\[
  C_i=\{c\in\SigmaSym_0:\alpha_c=a_i\}.
\]
Each \(C_i\) is finite.  Let
\[
  L_F=\bigcup_{a_i\in A_F}L_i.
\]
Then
\[
  U
  =
  (((L_F\cdot_{\lambda_1}C_1)\cdot_{\lambda_2}C_2)\cdots
    \cdot_{\lambda_m}C_m).       \tag{\(\dagger\)}
\]
Indeed, replacing every fresh constant \(\lambda_i\) by an original constant
from \(C_i\) preserves the state by the replacement lemma, so the right hand
side is contained in \(U\).  Conversely, if \(s\in U\), replace each original
constant \(c\) in \(s\) by the fresh constant \(\lambda_i\) such that
\(\alpha_c=a_i\).  The resulting \(\Sigma^+\)-term has the same root state and
therefore lies in \(L_F\), and \(s\) is recovered by the displayed iterated
complex product.  Since the right hand side of \((\dagger)\) is regular,
\(U\) is regular.
\end{proof}

\begin{theorem}[TW Theorem 8: recognizable iff regular]\label{thm:tw-recognizable-regular}
For every finite species, \(U\subseteq\Term{\SigmaSym}\) is recognizable iff it
is regular.
\end{theorem}

\begin{proof}
The forward implication is exactly the analysis theorem above.

For the converse, argue by induction on the construction of a regular set.
Finite sets are recognizable by Lemma~\ref{lem:tw-finite-recognizable}.
Recognizable sets are closed under finite union by Theorem~\ref{thm:tw-boolean}.
They are closed under complex product by
Lemma~\ref{lem:tw-complex-product}, and under complex closure by
Lemma~\ref{lem:tw-complex-closure}.  These are precisely the operations used to
generate regular sets, even after adding finitely many fresh constants.  Hence
every regular set is recognizable.
\end{proof}

\begin{lemma}[TW Lemma 10]\label{lem:tw-finite-recognizable}
Every finite subset of \(\Term{\SigmaSym}\) is recognizable.
\end{lemma}

\begin{proof}
It is enough to recognize singleton languages, because finite unions of
recognizable sets are recognizable by Theorem~\ref{thm:tw-boolean}.

Fix a term \(s\).  Let \(\mathrm{Sub}(s)\) be the finite set of subterms of
\(s\), and add one extra state \(\bot\).  Define a deterministic automaton
\(\Aut_s\) with state set
\[
  \mathrm{Sub}(s)\cup\{\bot\}.
\]
For a constant \(c\in\SigmaSym_0\), set
\[
  \alpha_c=
  \begin{cases}
    c, & c\in\mathrm{Sub}(s),\\
    \bot, & c\notin\mathrm{Sub}(s).
  \end{cases}
\]
For \(f\in\SigmaSym_n\) with \(n>0\), define
\[
  \alpha_f(q_1,\ldots,q_n)=
  \begin{cases}
    f(q_1,\ldots,q_n),
      & q_i\in\mathrm{Sub}(s)\text{ for all }i
        \text{ and }f(q_1,\ldots,q_n)\in\mathrm{Sub}(s),\\
    \bot, & \text{otherwise}.
  \end{cases}
\]
Here the expression \(f(q_1,\ldots,q_n)\) is used only in the first case, where
the \(q_i\)'s are genuine subterms.

We claim that for every input term \(t\),
\[
  h_{\Aut_s}(t)=
  \begin{cases}
    t, & t\in\mathrm{Sub}(s),\\
    \bot, & t\notin\mathrm{Sub}(s).
  \end{cases}        \tag{\(*\)}
\]
This is proved by structural induction on \(t\).  The constant case is exactly
the definition of \(\alpha_c\).  For \(t=f(t_1,\ldots,t_n)\), the induction
hypothesis says that each child state is either the child term \(t_i\), if
\(t_i\in\mathrm{Sub}(s)\), or \(\bot\), otherwise.  If any child is \(\bot\), the
transition returns \(\bot\).  If all children are subterms, the transition
returns \(f(t_1,\ldots,t_n)=t\) exactly when \(t\in\mathrm{Sub}(s)\), and returns
\(\bot\) otherwise.  Thus \((*)\) holds.

Taking the unique final state to be \(s\), we get
\[
  \bh_{\Aut_s}(\{s\})=\{s\}.
\]
Therefore every singleton is recognizable, and hence every finite set is
recognizable.
\end{proof}

\begin{lemma}[TW Lemma 11: closure under complex product]\label{lem:tw-complex-product}
If \(U,V\subseteq\Term{\SigmaSym}\) are recognizable, then
\[
  U\cdot_\lambda V
\]
is recognizable for every \(\lambda\in\SigmaSym_0\).
\end{lemma}

\begin{proof}
Let deterministic automata \(\Aut=(A,\alpha)\) and \(\mathcal B=(B,\beta)\)
recognize \(U\) and \(V\), with final sets \(A_F\) and \(B_F\).  We may assume
the state sets are disjoint, replacing them by tagged copies if necessary.

Construct a nondeterministic automaton \(\mathcal N\) with state set
\[
  A\cup B.
\]
The \(B\)-states simulate the automaton for \(V\).  The \(A\)-states simulate
the outer automaton for \(U\).  Whenever a subtree is accepted by the
\(V\)-automaton, \(\mathcal N\) may also treat the whole subtree as one
\(\lambda\)-leaf for the \(U\)-automaton.

For a constant \(c\in\SigmaSym_0\), put \(\beta_c\) into the set of possible
states.  If \(c\ne\lambda\), also put \(\alpha_c\) into the set of possible
states.  Finally, if \(\beta_c\in B_F\), also put \(\alpha_\lambda\) into the
set of possible states.  Thus an accepted \(V\)-constant can be collapsed to a
\(\lambda\)-hole for the outer computation.

For \(f\in\SigmaSym_n\), \(n>0\), the transition relation contains:
\begin{itemize}[leftmargin=*]
  \item the \(B\)-simulation transition
  \[
    (\beta_f(b_1,\ldots,b_n))
    \quad\text{from }(b_1,\ldots,b_n)\in B^n;
  \]
  \item the \(A\)-simulation transition
  \[
    \alpha_f(a_1,\ldots,a_n)
    \quad\text{from }(a_1,\ldots,a_n)\in A^n;
  \]
  \item the collapse transition to \(\alpha_\lambda\) from
  \((b_1,\ldots,b_n)\in B^n\) whenever
  \[
    \beta_f(b_1,\ldots,b_n)\in B_F.
  \]
\end{itemize}
Use \(A_F\) as the accepting set of \(\mathcal N\).

We prove two invariants by induction on the input term \(t\).
\[
  b\in h_{\mathcal N}(t)\cap B
  \quad\Longleftrightarrow\quad
  h_{\mathcal B}(t)=b.                       \tag{B}
\]
For constants this is just the definition of the initial \(B\)-state
\(\beta_c\).  For \(t=f(t_1,\ldots,t_n)\), a \(B\)-state can be produced only
by the \(B\)-simulation transition, so it has the form
\[
  \beta_f(b_1,\ldots,b_n)
\]
with \(b_i\in h_{\mathcal N}(t_i)\cap B\).  By the induction hypothesis,
\(b_i=h_{\mathcal B}(t_i)\), hence the produced state is exactly
\[
  \beta_f(h_{\mathcal B}(t_1),\ldots,h_{\mathcal B}(t_n))
  =
  h_{\mathcal B}(t).
\]
The converse direction is obtained by taking the unique \(B\)-states
\(h_{\mathcal B}(t_i)\) in the children and applying the \(B\)-simulation
transition.  Thus (B) holds.

The \(A\)-state invariant is:
\[
  a\in h_{\mathcal N}(t)\cap A
  \quad\Longleftrightarrow\quad
  \exists u\in\Term{\SigmaSym},\
    t\in\{u\}\cdot_\lambda V
    \text{ and } h_\Aut(u)=a.                \tag{A}
\]
For constants, there are two ways to produce an \(A\)-state.  If
\(c\ne\lambda\), the ordinary \(A\)-state \(\alpha_c\) is produced; the witness
is \(u=c\), and \(c\in\{c\}\cdot_\lambda V\).  If \(\beta_c\in B_F\), then
\(\mathcal N\) may produce \(\alpha_\lambda\); by (B), \(c\in V\), so the witness
is \(u=\lambda\), since \(c\in\{\lambda\}\cdot_\lambda V\).  Conversely, these
are the only possibilities for a constant input: if
\[
  c\in\{u\}\cdot_\lambda V,
\]
then either \(u=c\ne\lambda\), or \(u=\lambda\) and \(c\in V\).

For the inductive step, let \(t=f(t_1,\ldots,t_n)\).  If \(\mathcal N\) produces
an \(A\)-state by an \(A\)-simulation transition, then the child states have
witnesses \(u_i\) by the induction hypothesis.  Taking
\[
  u=f(u_1,\ldots,u_n)
\]
gives \(t\in\{u\}\cdot_\lambda V\), and the produced state is
\[
  \alpha_f(h_\Aut(u_1),\ldots,h_\Aut(u_n))=h_\Aut(u).
\]
If \(\mathcal N\) produces \(\alpha_\lambda\) by a collapse transition, then by
(B) the whole term \(t\) is accepted by \(\mathcal B\), so \(t\in V\); take
\(u=\lambda\).  This proves the left-to-right direction of (A).

For the converse direction of (A), suppose
\[
  t\in\{u\}\cdot_\lambda V
  \qquad\text{and}\qquad
  h_\Aut(u)=a.
\]
If \(u=\lambda\), then \(t\in V\), so \(h_{\mathcal B}(t)\in B_F\).  By (B), the
\(B\)-run of \(t\) is available in \(\mathcal N\), and the collapse rule produces
\[
  \alpha_\lambda=h_\Aut(\lambda)=a.
\]
If \(u=c\ne\lambda\) is a constant, then \(t=c\) and the initial \(A\)-state
\(\alpha_c=a\) is present.  Finally, if \(u=f(u_1,\ldots,u_n)\), then
\[
  t=f(t_1,\ldots,t_n)
\]
with \(t_i\in\{u_i\}\cdot_\lambda V\) for each \(i\).  By induction, each child
run contains the state \(h_\Aut(u_i)\).  The \(A\)-simulation transition then
produces
\[
  \alpha_f(h_\Aut(u_1),\ldots,h_\Aut(u_n))
  =
  h_\Aut(u)
  =
  a.
\]
This completes the proof of (A).

Now
\[
  t\in\bh_{\mathcal N}(A_F)
  \Longleftrightarrow
  \exists a\in A_F,\ a\in h_{\mathcal N}(t)
  \Longleftrightarrow
  \exists u\in U,\ t\in\{u\}\cdot_\lambda V
  \Longleftrightarrow
  t\in U\cdot_\lambda V.
\]
Therefore \(U\cdot_\lambda V\) is recognizable.
\end{proof}

\begin{lemma}[TW Lemma 12: closure under complex closure]\label{lem:tw-complex-closure}
If \(U\subseteq\Term{\SigmaSym}\) is recognizable, then \(U^{*_\lambda}\) is
recognizable.
\end{lemma}

\begin{proof}
Let \(\Aut=(A,\alpha)\) recognize \(U\), with final set \(A_F\).  We use a
tagged copy of \(A\), written
\[
  \widehat A=\{\widehat a:a\in A\},
\]
and one fresh state \(q_\lambda\).  The state \(q_\lambda\) means ``this subtree
has already been proved to lie in the closure and may be used as a
\(\lambda\)-replacement.''  Define
\[
  \mathrm{val}(\widehat a)=a,
  \qquad
  \mathrm{val}(q_\lambda)=\alpha_\lambda.
\]
Construct a nondeterministic automaton \(\mathcal N\) with state set
\[
  \widehat A\cup\{q_\lambda\}.
\]

For constants \(c\in\SigmaSym_0\), the initial states are:
\[
  q_\lambda \quad\text{if } c=\lambda,
\]
\[
  \widehat{\alpha_c} \quad\text{if } c\ne\lambda,
\]
and also
\[
  q_\lambda \quad\text{if } c\ne\lambda\text{ and }\alpha_c\in A_F.
\]
The last clause says that an input constant already accepted as a member of
\(U\) may be collapsed to a \(\lambda\)-replacement.  The first clause supplies
the base element \(\lambda\in U^{*_\lambda}\).  We deliberately do not use
\(\widehat{\alpha_\lambda}\) as the accepting marker; \(q_\lambda\) is fresh, so
no unrelated term can be accepted merely because its ordinary automaton state is
\(\alpha_\lambda\).

For \(f\in\SigmaSym_n\), \(n>0\), and child states \(r_1,\ldots,r_n\), let
\[
  a=\alpha_f(\mathrm{val}(r_1),\ldots,\mathrm{val}(r_n)).
\]
The transition relation allows output \(\widehat a\).  If \(a\in A_F\), it also
allows output \(q_\lambda\).  The final set is the singleton
\[
  \{q_\lambda\}.
\]

Let \(X_0=\{\lambda\}\) and
\[
  X_{r+1}=X_r\cup(U\cdot_\lambda X_r),
  \qquad
  U^{*_\lambda}=\bigcup_r X_r.
\]
The sequence is monotone, since \(X_r\subseteq X_{r+1}\) for every \(r\).
We use the following elementary consequence of the definition: if
\[
  C=U^{*_\lambda},
\]
then
\[
  U\cdot_\lambda C\subseteq C.              \tag{1}
\]
Indeed, if \(t\in U\cdot_\lambda C\), then \(t\) is obtained from some
\(u\in U\) by replacing the finitely many \(\lambda\)-occurrences of \(u\) by
terms \(s_1,\ldots,s_m\in C\).  Choose stages \(r_i\) with \(s_i\in X_{r_i}\),
and let \(r=\max_i r_i\) (with \(r=0\) if there are no such occurrences).  Then
all \(s_i\in X_r\), so
\[
  t\in U\cdot_\lambda X_r\subseteq X_{r+1}\subseteq C.
\]

We now prove, by induction on \(r\), that
\[
  t\in X_r \quad\Longrightarrow\quad q_\lambda\in h_{\mathcal N}(t).      \tag{3}
\]
For \(r=0\), \(t=\lambda\), and \(q_\lambda\) is an initial state at
\(\lambda\).  Suppose the claim is known for \(X_r\).  We first derive the
auxiliary assertion
\[
  u\ne\lambda\text{ and }t\in\{u\}\cdot_\lambda X_r
  \quad\Longrightarrow\quad
  \widehat{h_\Aut(u)}\in h_{\mathcal N}(t),        \tag{2}
\]
by structural induction on \(u\).

If \(u=c\ne\lambda\) is a constant, then \(t=c\), and the initial state
\(\widehat{\alpha_c}\) is present.  If
\[
  u=f(u_1,\ldots,u_n),
\]
then \(t=f(t_1,\ldots,t_n)\), where
\[
  t_i\in\{u_i\}\cdot_\lambda X_r
\]
for each \(i\).  For a child \(u_i=\lambda\), this says \(t_i\in X_r\), so the
induction hypothesis on the stage gives
\[
  q_\lambda\in h_{\mathcal N}(t_i).
\]
For a child \(u_i\ne\lambda\), the structural induction hypothesis gives
\[
  \widehat{h_\Aut(u_i)}\in h_{\mathcal N}(t_i).
\]
In both cases we have chosen a child state \(r_i\) with
\[
  \mathrm{val}(r_i)=h_\Aut(u_i).
\]
The transition for \(f\) therefore produces
\[
  \widehat{\alpha_f(h_\Aut(u_1),\ldots,h_\Aut(u_n))}
  =
  \widehat{h_\Aut(u)}.
\]
This proves (2) under the induction hypothesis for \(X_r\).

Now let \(t\in X_{r+1}\).
If \(t\in X_r\), we are done.  Otherwise
\[
  t\in U\cdot_\lambda X_r.
\]
Choose \(u\in U\) with \(t\in\{u\}\cdot_\lambda X_r\).  If \(u=\lambda\), then
\(t\in X_r\), so the induction hypothesis applies.  If \(u\ne\lambda\), (2)
gives \(\widehat{h_\Aut(u)}\in h_{\mathcal N}(t)\).  Since \(u\in U\), we have
\[
  h_\Aut(u)\in A_F.
\]
If \(u=c\ne\lambda\) is a constant, then \(t=c\), and the extra initial-state
clause for final constants gives \(q_\lambda\).  If \(u\) has positive arity,
the last transition used in the proof of (2) computes the value \(h_\Aut(u)\);
because this value is final, the same child states also allow the output
\(q_\lambda\).  This proves (3).

It remains to prove the converse implication
\[
  q_\lambda\in h_{\mathcal N}(t)
  \quad\Longrightarrow\quad
  t\in U^{*_\lambda}.                         \tag{4}
\]
The reverse implication follows already from (3), since
\(U^{*_\lambda}=\bigcup_r X_r\).

To prove (4), we prove simultaneously, by structural
induction on \(t\), the following two assertions:
\[
  q_\lambda\in h_{\mathcal N}(t)\Rightarrow t\in C,             \tag{Q}
\]
and
\[
  \widehat a\in h_{\mathcal N}(t)
  \Rightarrow
  \exists u\ne\lambda,\ t\in\{u\}\cdot_\lambda C
    \text{ and } h_\Aut(u)=a.                                  \tag{A}
\]
For a constant \(c\), assertion (Q) holds because \(q_\lambda\) is initial only
when either \(c=\lambda\), hence \(c\in X_0\), or \(c\ne\lambda\) and
\(\alpha_c\in A_F\), hence \(c\in U\cdot_\lambda X_0\subseteq C\).  Assertion
(A) holds because the only tagged initial state is \(\widehat{\alpha_c}\) with
\(c\ne\lambda\), witnessed by \(u=c\).

Let \(t=f(t_1,\ldots,t_n)\).  If \(\widehat a\in h_{\mathcal N}(t)\), then it is
produced from child states \(r_i\in h_{\mathcal N}(t_i)\).  If
\(r_i=q_\lambda\), use (Q) for \(t_i\) and set \(u_i=\lambda\).  If
\(r_i=\widehat{a_i}\), use (A) for \(t_i\) to choose \(u_i\ne\lambda\) with
\[
  t_i\in\{u_i\}\cdot_\lambda C
  \quad\text{and}\quad
  h_\Aut(u_i)=a_i.
\]
Set \(u=f(u_1,\ldots,u_n)\).  Then \(u\ne\lambda\),
\[
  t\in\{u\}\cdot_\lambda C,
\]
by the defining, coordinatewise clause for the complex product, and, by the
definition of \(\mathrm{val}\),
\[
  h_\Aut(u)=\alpha_f(\mathrm{val}(r_1),\ldots,\mathrm{val}(r_n))=a.
\]
This proves (A) for \(t\).

If \(q_\lambda\in h_{\mathcal N}(t)\), then it is produced by a collapse rule
from some child states \(r_i\), and the corresponding value
\[
  a=\alpha_f(\mathrm{val}(r_1),\ldots,\mathrm{val}(r_n))
\]
lies in \(A_F\).  By the previous paragraph, applied to the tagged state
\(\widehat a\) produced by the same child states, there is \(u\ne\lambda\) with
\[
  t\in\{u\}\cdot_\lambda C
  \quad\text{and}\quad
  h_\Aut(u)=a\in A_F.
\]
Thus \(u\in U\), so \(t\in U\cdot_\lambda C\subseteq C\) by (1).  This proves
(Q), and hence (4).

Therefore
\[
  q_\lambda\in h_{\mathcal N}(t)
  \quad\Longleftrightarrow\quad
  t\in U^{*_\lambda},
\]
so \(\mathcal N\), with final state \(q_\lambda\), recognizes
\(U^{*_\lambda}\).
\end{proof}

\begin{definition}[Truncation, TW \S3]
For \(U,V\subseteq\Term{\SigmaSym}\), define
\[
  U/_{\lambda}V
  =
  \{t: ( \{t\}\cdot_\lambda V )\cap U\neq\emptyset\}.
\]
Intuitively, \(t\) is obtained from some term of \(U\) by replacing selected,
non-overlapping subterms from \(V\) by \(\lambda\), with every occurrence of
\(\lambda\) covered by one of the removed subterms.
\end{definition}

\begin{lemma}[TW Lemma 13: closure under truncation]\label{lem:tw-truncation}
If \(U\subseteq\Term{\SigmaSym}\) is recognizable and
\(V\subseteq\Term{\SigmaSym}\) is arbitrary, then \(U/_{\lambda}V\) is
recognizable.
\end{lemma}

\begin{proof}
Let \(\Aut=(A,\alpha)\) recognize \(U\), with final set \(A_F\), so
\[
  U=\bh_\Aut(A_F).
\]
Although \(V\) is arbitrary, we only need to remember which states of
\(\Aut\) are reached by terms of \(V\).  Put
\[
  A_V=\{h_\Aut(v):v\in V\}\subseteq A.
\]
This is a finite set because \(A\) is finite.

Construct a nondeterministic automaton \(\mathcal N=(A,\rho)\) on the same state
set \(A\).  For constants \(c\in\SigmaSym_0\), define the initial-state
relation by
\[
  \rho_c=
  \begin{cases}
    \{\alpha_c\}, & c\ne\lambda,\\
    A_V,          & c=\lambda.
  \end{cases}
\]
For a positive-arity symbol \(f\in\SigmaSym_n\), \(n>0\), let \(\rho_f\) be the
graph of the original deterministic transition:
\[
  \rho_f(a_1,\ldots,a_n,a)
  \quad\Longleftrightarrow\quad
  \alpha_f(a_1,\ldots,a_n)=a.
\]
Thus \(\mathcal N\) behaves exactly like \(\Aut\) except at a visible
\(\lambda\)-leaf, where it may guess the state of some term from \(V\).  We use
the same accepting set \(A_F\).

We prove two statements by structural induction on the visible term \(t\):
\[
  s\in\{t\}\cdot_\lambda V
  \quad\Longrightarrow\quad
  h_\Aut(s)\in h_{\mathcal N}(t),             \tag{I}
\]
and
\[
  a\in h_{\mathcal N}(t)
  \quad\Longrightarrow\quad
  \exists s\in\{t\}\cdot_\lambda V,\ h_\Aut(s)=a.   \tag{II}
\]
Statement (I) says that every genuine expansion of \(t\) by \(V\)-terms is
simulated by the nondeterministic run on the truncated term \(t\).  Statement
(II) says that every nondeterministic choice made by \(\mathcal N\) is realized
by some such expansion.

First let \(t=c\) be a constant.  If \(c\ne\lambda\), then
\[
  \{c\}\cdot_\lambda V=\{c\}
\]
and
\[
  h_{\mathcal N}(c)=\{\alpha_c\}.
\]
Hence (I) and (II) both hold with the unique witness \(s=c\).  If
\(c=\lambda\), then
\[
  \{\lambda\}\cdot_\lambda V=V
  \qquad\text{and}\qquad
  h_{\mathcal N}(\lambda)=A_V.
\]
For (I), if \(s\in V\), then \(h_\Aut(s)\in A_V=h_{\mathcal N}(\lambda)\).
For (II), if \(a\in h_{\mathcal N}(\lambda)=A_V\), then by definition of
\(A_V\) there is some \(s\in V=\{\lambda\}\cdot_\lambda V\) with
\[
  h_\Aut(s)=a.
\]

Now let
\[
  t=f(t_1,\ldots,t_n)
\]
with \(n>0\), and assume (I) and (II) for \(t_1,\ldots,t_n\).

For (I), suppose
\[
  s\in\{t\}\cdot_\lambda V.
\]
By the recursive definition of complex product, \(s\) has the form
\[
  s=f(s_1,\ldots,s_n)
\]
with
\[
  s_i\in\{t_i\}\cdot_\lambda V
  \qquad (i=1,\ldots,n).
\]
By the induction hypothesis for (I),
\[
  h_\Aut(s_i)\in h_{\mathcal N}(t_i)
  \qquad (i=1,\ldots,n).
\]
Since \(\rho_f\) is the graph of \(\alpha_f\), these child states produce the
root state
\[
  \alpha_f(h_\Aut(s_1),\ldots,h_\Aut(s_n))
  =
  h_\Aut(s)
\]
in the nondeterministic run of \(t\).  Therefore
\[
  h_\Aut(s)\in h_{\mathcal N}(t).
\]

For (II), suppose \(a\in h_{\mathcal N}(t)\).  By the definition of the
nondeterministic run, there are states
\[
  a_i\in h_{\mathcal N}(t_i)
  \qquad (i=1,\ldots,n)
\]
such that
\[
  \rho_f(a_1,\ldots,a_n,a).
\]
Because \(\rho_f\) is the graph of \(\alpha_f\), this means
\[
  a=\alpha_f(a_1,\ldots,a_n).
\]
By the induction hypothesis for (II), for each \(i\) choose
\[
  s_i\in\{t_i\}\cdot_\lambda V
  \qquad\text{with}\qquad
  h_\Aut(s_i)=a_i.
\]
Set
\[
  s=f(s_1,\ldots,s_n).
\]
Then, again by the recursive definition of complex product,
\[
  s\in\{t\}\cdot_\lambda V,
\]
and
\[
  h_\Aut(s)
  =
  \alpha_f(h_\Aut(s_1),\ldots,h_\Aut(s_n))
  =
  \alpha_f(a_1,\ldots,a_n)
  =
  a.
\]
This proves (II).
Therefore the simultaneous structural induction establishes (I) and (II) for
all terms \(t\).

It remains to identify the accepted language of \(\mathcal N\).  If
\[
  t\in U/_{\lambda}V,
\]
then by definition there is some
\[
  s\in(\{t\}\cdot_\lambda V)\cap U.
\]
Since \(s\in U=\bh_\Aut(A_F)\), we have \(h_\Aut(s)\in A_F\).  By (I),
\[
  h_\Aut(s)\in h_{\mathcal N}(t),
\]
so \(h_{\mathcal N}(t)\cap A_F\neq\emptyset\), and \(t\) is accepted by
\(\mathcal N\).

Conversely, suppose \(t\) is accepted by \(\mathcal N\).  Then there is some
\[
  a\in h_{\mathcal N}(t)\cap A_F.
\]
By (II), choose \(s\in\{t\}\cdot_\lambda V\) with \(h_\Aut(s)=a\).  Since
\(a\in A_F\), the term \(s\) lies in \(U\).  Hence
\[
  s\in(\{t\}\cdot_\lambda V)\cap U,
\]
so \((\{t\}\cdot_\lambda V)\cap U\neq\emptyset\), and therefore
\[
  t\in U/_{\lambda}V.
\]
Thus \(\bh_{\mathcal N}(A_F)=U/_{\lambda}V\), so \(U/_{\lambda}V\) is
recognizable.
\end{proof}

\section{Weak MSO Over the Binary Successor Tree}

\TW{} apply the automata theory to the weak monadic second-order theory of
multiple successors.  For Courcelle, this is the important logic-to-automata
part of the paper.

\begin{definition}[The tree \(N_2\)]
Let \(N_2=\{1,2\}^*\), with empty word \(\epsilon\), and successor functions
\[
  r_i(w)=wi \qquad (i=1,2).
\]
The weak MSO language has first-order variables ranging over \(N_2\), set
variables ranging over finite subsets of \(N_2\), equality, membership, and
binary successor predicates \(R_i(x,y)\), interpreted as \(r_i(x)=y\).
\end{definition}

\begin{theorem}[TW Theorem 14, Doner]\label{thm:tw-doner}
The weak monadic second-order theory of multiple successors is decidable.
\end{theorem}

\noindent
\TW{} state this theorem before developing the machinery used in its proof.  We
follow that order: the proof is given at the end of this section, after the
set-only translation, the finite-set encoding, and Theorem~\ref{thm:tw-definable-recognizable}.

\section{Eliminating First-Order Variables}\label{sec:tw-eliminate-first-order}

\begin{definition}[Set-only translation]
Replace each first-order variable \(x\) by a set variable \(X_x\), constrained to
be a singleton.  A convenient cleaned definition of singleton, using only
\(\subseteq\), is:
\[
  \mathrm{Empty}(A) \;:\!\!\Longleftrightarrow\; \forall B,\ A\subseteq B,
\]
\[
  \mathrm{Sing}(A) \;:\!\!\Longleftrightarrow\;
  \neg\mathrm{Empty}(A)\ \wedge\
  \forall B,\ B\subseteq A \to (\mathrm{Empty}(B)\vee A\subseteq B).
\]
The atomic translation is:
\[
  (x=y)^* \;=\; X_x\subseteq X_y\wedge X_y\subseteq X_x\wedge
  \mathrm{Sing}(X_x)\wedge
  \mathrm{Sing}(X_y),
\]
\[
  (x\in A)^* \;=\; X_x\subseteq A\wedge\mathrm{Sing}(X_x),
\]
\[
  (R_i(x,y))^* \;=\; \widehat R_i(X_x,X_y),
\]
where \(\widehat R_i\) is the successor relation restricted to singleton sets.
The translation commutes with Boolean connectives and sends
\[
  (\exists x\,\varphi)^*
  =
  \exists X_x\,(\mathrm{Sing}(X_x)\wedge\varphi^*),
  \qquad
  (\exists A\,\varphi)^*=\exists A\,\varphi^*.
\]
\end{definition}

\begin{remark}
The PDF gives an equivalent formula for \(\mathrm{Sing}\), but the OCR in that
line is especially noisy.  The formula above is the same idea in a form better
suited to formalization.
\end{remark}

\medskip
\noindent\textbf{Verification of the translation.}
There are two facts to check.

First, the auxiliary predicates have the intended meanings.  For any finite
set \(A\subseteq N_2\),
\[
  \mathrm{Empty}(A)
  \quad\Longleftrightarrow\quad
  A=\emptyset.
\]
Indeed, if \(A\subseteq B\) for every \(B\), take \(B=\emptyset\).  Conversely,
\(\emptyset\subseteq B\) for every \(B\).  Also
\[
  \mathrm{Sing}(A)
  \quad\Longleftrightarrow\quad
  \exists w\in N_2,\ A=\{w\}.
\]
If \(\mathrm{Sing}(A)\) holds, then \(A\) is nonempty, so choose \(w\in A\).
The subset \(\{w\}\subseteq A\) is not empty, so the defining condition for
\(\mathrm{Sing}\) gives \(A\subseteq\{w\}\), hence \(A=\{w\}\).  Conversely, if
\(A=\{w\}\), then \(A\) is not empty, and every subset of \(A\) is either
\(\emptyset\) or \(A\) itself.

Second, the translation preserves truth.  Fix a formula
\[
  \varphi(x_1,\ldots,x_r,A_1,\ldots,A_m)
\]
of the original weak MSO language.  Given an assignment
\[
  x_j\mapsto w_j\in N_2,\qquad A_\ell\mapsto B_\ell\in\powfin(N_2),
\]
let the translated assignment send
\[
  X_{x_j}\mapsto \{w_j\},\qquad A_\ell\mapsto B_\ell.
\]
Then
\[
  (w_1,\ldots,w_r,B_1,\ldots,B_m)\models\varphi
  \quad\Longleftrightarrow\quad
  (\{w_1\},\ldots,\{w_r\},B_1,\ldots,B_m)\models\varphi^*.       \tag{1}
\]
This is proved by structural induction on \(\varphi\).  For atomic formulas,
\[
  x=y
\]
is equivalent to
\[
  \{w_x\}\subseteq\{w_y\}\wedge \{w_y\}\subseteq\{w_x\},
\]
membership \(x\in A\) is equivalent to \(\{w_x\}\subseteq A\), and
\[
  R_i(x,y)
\]
is equivalent to \(\widehat R_i(\{w_x\},\{w_y\})\) by definition of
\(\widehat R_i\).  Boolean connectives are immediate from the induction
hypothesis.  For \(\exists x\,\psi\), a witness \(w\in N_2\) corresponds to the
singleton witness \(\{w\}\) satisfying \(\mathrm{Sing}\), and conversely any
set satisfying \(\mathrm{Sing}\) is \(\{w\}\) for a unique \(w\).  Existential
set quantifiers are unchanged.  In particular, if \(S\) is a sentence, then
\[
  S\text{ is true}
  \quad\Longleftrightarrow\quad
  S^*\text{ is true}.
\]

\section{Encoding Finite Subsets as Terms}

\begin{definition}[Single-track encoding]
Let \(\Sigma_1\) have one nullary symbol \(\lambda\) and two binary symbols
\(0,1\).  Define
\[
  c:\Term{\Sigma_1}\to \powfin(N_2)
\]
by
\[
  c(\lambda)=\emptyset,
\]
\[
  c(0(t_1,t_2))=1c(t_1)\cup 2c(t_2),
\]
\[
  c(1(t_1,t_2))=\{\epsilon\}\cup 1c(t_1)\cup 2c(t_2),
\]
where \(ic(t)=\{iw:w\in c(t)\}\).  Thus a node labelled \(1\) marks the
corresponding position as present in the finite subset of \(N_2\), and a node
labelled \(0\) leaves it absent.
\end{definition}

\begin{definition}[Canonical encoding]
The map \(c\) is onto but not one-to-one, since all-zero subtrees carry no set
information.  For a finite set \(A\subseteq N_2\), let \(e(A)\) be the smallest
term in \(c^{-1}(A)\), equivalently the unique representative with no subterm of
the form \(0(\lambda,\lambda)\).
\end{definition}

\begin{lemma}[TW Lemma 15]\label{lem:tw-encoding-one}
The set
\[
  U_1=\{e(A):A\in\powfin(N_2)\}
\]
of canonical encodings of finite subsets of \(N_2\) is recognizable.
\end{lemma}

\begin{proof}
We prove that \(U_1\) is regular.  By
Theorem~\ref{thm:tw-recognizable-regular}, this implies recognizability.

Add one fresh constant \(\delta\), used as a temporary hole.  Let \(P_1\) be the
finite set of one-step patterns
\[
  P_1=
  \{f(x_1,x_2):
      f\in\{0,1\},\ x_1,x_2\in\{\lambda,\delta\},
      \text{ and } f(x_1,x_2)\ne 0(\lambda,\lambda)\}.
\]
Thus \(P_1\) contains all one-node \(0/1\)-terms with children
\(\lambda\) or \(\delta\), except for the forbidden redundant subtree
\(0(\lambda,\lambda)\).  Put
\[
  W_1=P_1^{*_\delta}.
\]
Equivalently, if
\[
  X_0=\{\delta\},\qquad
  X_{r+1}=X_r\cup(P_1\cdot_\delta X_r),
\]
then \(W_1=\bigcup_r X_r\).  We claim that
\[
  U_1=(W_1\cdot_\delta\emptyset)\cup\{\lambda\}.        \tag{1}
\]

First let \(t\in W_1\cdot_\delta\emptyset\).  Since no term can be substituted
for \(\delta\) from the empty set, this means that \(t\in W_1\) and \(t\) has no
\(\delta\)-occurrence.  We show that no term of \(W_1\) has
\(0(\lambda,\lambda)\) as a subterm.  This is by induction on the closure stage
of \(W_1\).  The base term \(\delta\) has no such subterm.  In the step, a term
is obtained from a pattern in \(P_1\) by replacing its \(\delta\)-leaves by
previous-stage terms.  The root pattern is not \(0(\lambda,\lambda)\), by the
definition of \(P_1\), and the inserted terms have no forbidden subterm by the
induction hypothesis.  Hence \(t\) has no forbidden subterm.  Since \(t\) also
contains no \(\delta\), it is a genuine \(\Sigma_1\)-term with no subterm
\(0(\lambda,\lambda)\), so by the characterization of canonical encodings,
\[
  t\in U_1.
\]
Also \(\lambda=e(\emptyset)\), so \(\lambda\in U_1\).  This proves the
right-to-left inclusion in (1).

Conversely, let \(t\in U_1\).  If \(t=\lambda\), then \(t\) belongs to the
right hand side of (1).  Suppose \(t\ne\lambda\).  We prove by structural
induction that every non-\(\lambda\) canonical term lies in \(W_1\).  Write
\[
  t=f(t_1,t_2),\qquad f\in\{0,1\}.
\]
Since \(t\) is canonical, it is not the forbidden term \(0(\lambda,\lambda)\),
and each non-\(\lambda\) child \(t_i\) is again canonical.  For each child set
\[
  y_i=
  \begin{cases}
    \lambda, & t_i=\lambda,\\
    \delta, & t_i\ne\lambda.
  \end{cases}
\]
Then
\[
  p=f(y_1,y_2)\in P_1.
\]
For every child with \(t_i\ne\lambda\), the induction hypothesis gives
\[
  t_i\in W_1.
\]
Using \(p\), and replacing its \(\delta\)-occurrences by the corresponding
non-\(\lambda\) children, gives
\[
  t\in P_1\cdot_\delta W_1\subseteq W_1.
\]
The last inclusion follows from the definition of complex closure: if the
finitely many inserted terms lie in stages \(X_{r_i}\), choose
\(r=\max_i r_i\), and then the resulting term lies in \(X_{r+1}\).  Since the
original canonical term \(t\) contains no \(\delta\), we also have
\[
  t\in W_1\cdot_\delta\emptyset.
\]
This proves (1).

The right hand side of (1) is regular: \(P_1\), \(\emptyset\), and
\(\{\lambda\}\) are finite, and regular sets are closed under complex closure,
complex product, and finite union.  Therefore \(U_1\) is regular, hence
recognizable.
\end{proof}

\begin{definition}[Multi-track encoding]
For \(n\) set variables, use the species \(\Sigma_n\) with one nullary symbol
\(\lambda\) and binary symbols \(\{0,1\}^n\).  The projection
\[
  \pi_i:\{0,1\}^n\to\{0,1\}
\]
selects the \(i\)-th track and extends to terms.  Define
\[
  c_n(t)=\bigl(c(\pi_1(t)),\ldots,c(\pi_n(t))\bigr).
\]
For \(\bar A=(A_1,\ldots,A_n)\), \(e(\bar A)\) is the smallest term in
\(c_n^{-1}(\bar A)\), equivalently the representative with no subterm
\[
  (0,\ldots,0)(\lambda,\lambda).
\]
\end{definition}

\begin{lemma}[TW Lemma 16]\label{lem:tw-encoding-n}
The set
\[
  U_n=\{e(A_1,\ldots,A_n):(A_i\in\powfin(N_2))\}
\]
of canonical encodings of \(n\)-tuples of finite subsets of \(N_2\) is
recognizable.
\end{lemma}

\begin{proof}
The proof is the same as Lemma~\ref{lem:tw-encoding-one}, with the all-zero
\(n\)-track label replacing the one-track label \(0\).  We spell it out to make
the induction explicit.

Let
\[
  \mathbf 0=(0,\ldots,0)\in\{0,1\}^n,
\]
and add a fresh constant \(\delta\).  Define the finite pattern set
\[
  P_n=
  \{\ell(x_1,x_2):
      \ell\in\{0,1\}^n,\ x_1,x_2\in\{\lambda,\delta\},
      \text{ and } \ell(x_1,x_2)\ne
        \mathbf 0(\lambda,\lambda)\}.
\]
Put
\[
  W_n=P_n^{*_\delta}.
\]
We claim that
\[
  U_n=(W_n\cdot_\delta\emptyset)\cup\{\lambda\}.       \tag{2}
\]

For the inclusion from right to left, the same closure-stage induction as above
shows that no term of \(W_n\) contains the forbidden subterm
\[
  \mathbf 0(\lambda,\lambda).
\]
Indeed, the base term \(\delta\) has no such subterm; in the induction step, the
root pattern belongs to \(P_n\), so it is not the forbidden all-zero leaf, and
the inserted terms satisfy the induction hypothesis.  Thus any term in
\(W_n\cdot_\delta\emptyset\) is a genuine \(\Sigma_n\)-term with no
\(\delta\)-occurrence and no forbidden all-zero leaf.  By the characterization of
canonical \(n\)-track encodings, it belongs to \(U_n\).  The term \(\lambda\) is
the encoding of the empty \(n\)-tuple
\[
  (\emptyset,\ldots,\emptyset),
\]
so \(\lambda\in U_n\).

For the converse inclusion, let \(t\in U_n\).  If \(t=\lambda\), then \(t\) is
in the right hand side of (2).  Otherwise write
\[
  t=\ell(t_1,t_2),\qquad \ell\in\{0,1\}^n.
\]
We prove, by structural induction, that every non-\(\lambda\) canonical
\(n\)-track term lies in \(W_n\).  Since \(t\) is canonical, it is not
\[
  \mathbf 0(\lambda,\lambda),
\]
and each non-\(\lambda\) child \(t_i\) is canonical.  Define
\[
  y_i=
  \begin{cases}
    \lambda, & t_i=\lambda,\\
    \delta, & t_i\ne\lambda.
  \end{cases}
\]
Then
\[
  p=\ell(y_1,y_2)\in P_n.
\]
For every non-\(\lambda\) child, the induction hypothesis gives
\[
  t_i\in W_n.
\]
Replacing the \(\delta\)-occurrences of \(p\) by those children yields
\[
  t\in P_n\cdot_\delta W_n\subseteq W_n,
\]
where the last inclusion follows from the stage definition of
\(\delta\)-closure, just as in Lemma~\ref{lem:tw-encoding-one}.  Since \(t\) has
no \(\delta\)-occurrence, \(t\in W_n\cdot_\delta\emptyset\).  This proves (2).

The right hand side of (2) is regular, because it is built from finite sets by
complex closure, complex product, and finite union.  Hence \(U_n\) is regular
and therefore recognizable by Theorem~\ref{thm:tw-recognizable-regular}.
\end{proof}

\section{Definable Relations Are Recognizable}

\begin{definition}[Recognizable relation]
A relation \(R\subseteq(\powfin(N_2))^n\) is recognizable if
\[
  e(R)=\{e(\bar A):\bar A\in R\}
  \subseteq\Term{\Sigma_n}
\]
is recognizable.
\end{definition}

\begin{theorem}[TW Theorem 17]\label{thm:tw-definable-recognizable}
Every relation definable in the set-only weak MSO language over \(N_2\) is
recognizable.
\end{theorem}

\noindent
As in \TW{}, the proof is obtained from the two lemmas below: atomic formulas
are recognizable, and recognizable relations are closed under the logical
operations.  The proof is given after Lemma~\ref{lem:tw-logical-closure}.

\begin{lemma}[TW Lemma 18: atomic formulas]\label{lem:tw-atomic}
The relations on \(\powfin(N_2)\) defined by atomic formulas in the set-only
language are recognizable.
\end{lemma}

\begin{proof}
The atomic relations are subset \(A\subseteq B\) and singleton-successor
\(\widehat R_i(A,B)\).

For subset, work over the two-track alphabet \(\Sigma_2\).  Let \(N_{10}\) be
the set of all terms over \(\Sigma_2\) in which the binary label \((1,0)\) never
occurs.  This set is recognizable by the two-state deterministic automaton with
states
\[
  \mathrm{ok},\mathrm{bad}.
\]
The constant \(\lambda\) has state \(\mathrm{ok}\).  At a binary node with label
\(\ell\in\{0,1\}^2\), the transition returns \(\mathrm{bad}\) if
\(\ell=(1,0)\) or if one of the children is already \(\mathrm{bad}\); otherwise
it returns \(\mathrm{ok}\).  The unique accepting state is \(\mathrm{ok}\).

For a canonical two-track encoding \(e(A,B)\), the label at a position \(w\) is
\[
  (1,0)
  \quad\Longleftrightarrow\quad
  w\in A \text{ and } w\notin B.
\]
Hence
\[
  A\subseteq B
  \quad\Longleftrightarrow\quad
  e(A,B)\in N_{10}.
\]
Therefore
\[
  e(\subseteq)=N_{10}\cap U_2,
\]
which is recognizable by Boolean closure and Lemma~\ref{lem:tw-encoding-n}.

Now consider \(\widehat R_i(A,B)\).  This relation says that \(A\) and \(B\) are
singletons, say
\[
  A=\{w\},\qquad B=\{wi\}.
\]
\TW{} describe this relation by a regular expression.  We spell out the
construction because it is a useful induction pattern.

Introduce a fresh constant \(\delta\), used only as a temporary hole.  Let
\[
  H=\{(0,0)(\delta,\lambda),\ (0,0)(\lambda,\delta)\}.
\]
Thus \(H^{*_\delta}\) is the set of chains of \((0,0)\)-nodes with one
\(\delta\)-hole at the bottom and \(\lambda\)-side branches.  Define the two base
terms
\[
  B_1=(1,0)((0,1)(\lambda,\lambda),\lambda),
\]
\[
  B_2=(1,0)(\lambda,(0,1)(\lambda,\lambda)).
\]
Finally set
\[
  S_i=H^{*_\delta}\cdot_\delta\{B_i\}.
\]
The set \(H\) is finite, hence recognizable; by Lemmas
\ref{lem:tw-complex-closure} and \ref{lem:tw-complex-product}, each \(S_i\) is
recognizable.

We claim that
\[
  S_i=e(\widehat R_i).
\]
For the inclusion \(S_i\subseteq e(\widehat R_i)\), induct on the stage in the
\(\delta\)-closure defining \(H^{*_\delta}\).  At stage \(0\), substituting
\(B_i\) for the hole gives the encoding of
\[
  (\{\epsilon\},\{i\}).
\]
At the next stages, the only new operations are
\[
  t\mapsto (0,0)(t,\lambda)
  \qquad\text{and}\qquad
  t\mapsto (0,0)(\lambda,t).
\]
The first operation prefixes both singleton positions by \(1\), and the second
prefixes both singleton positions by \(2\).  Therefore, if the old pair is
\((\{w\},\{wi\})\), the new pair is either
\[
  (\{1w\},\{1wi\})
  \qquad\text{or}\qquad
  (\{2w\},\{2wi\}),
\]
so the successor relation is preserved.  The side branch is \(\lambda\), and
the active child is not an all-zero redundant subtree, so the resulting term is
canonical.

Conversely, suppose \(e(A,B)\) satisfies \(\widehat R_i\), say
\[
  A=\{w\},\qquad B=\{wi\}.
\]
Induct on the length of \(w\).  If \(w=\epsilon\), then the canonical encoding is
exactly \(B_i\), hence belongs to \(S_i\).  If \(w=1u\), then the canonical
encoding has the form
\[
  (0,0)(t,\lambda),
\]
where \(t=e(\{u\},\{ui\})\).  By induction \(t\in S_i\), and the left wrapper
\((0,0)(\delta,\lambda)\in H\) shows that
\((0,0)(t,\lambda)\in S_i\).  The case \(w=2u\) is identical, using the right
wrapper \((0,0)(\lambda,\delta)\).  Thus \(e(\widehat R_i)=S_i\), and
\(\widehat R_i\) is recognizable.
\end{proof}

\begin{lemma}[TW Lemma 19: logical closure]\label{lem:tw-logical-closure}
If \(R\) and \(S\) are recognizable relations, then the following are
recognizable:
\[
  R(\bar A,\bar B)\wedge S(\bar B,\bar C),\qquad
  \neg R(\bar A),\qquad
  \exists A_i\,R(A_1,\ldots,A_m,\bar B).
\]
\end{lemma}

\begin{proof}
For \(k\ge 0\), let \(\mathbf 0_k\in\{0,1\}^k\) be the all-zero tuple; when
\(k=0\), this is the unique empty tuple.  Write
\[
  z_k=\mathbf 0_k(\lambda,\lambda)\in\Term{\Sigma_k}
\]
for the one-node all-zero binary term.  For \(k=0\), \(\Sigma_0\) is understood
as the zero-track species with the nullary symbol \(\lambda\) and the single
binary label \(\mathbf 0_0\).  Put
\[
  Z_k=\{z_k\}^{*_\lambda}.
\]
Thus \(Z_k\) is the set of all all-zero \(k\)-track terms, including
\(\lambda\).  This is proved by induction on the closure stage in one direction
and by structural induction on all-zero terms in the other.  In particular,
substituting a term from \(Z_k\) into a \(\lambda\)-leaf never changes the
encoded \(k\)-tuple of finite subsets.  The set \(Z_k\) is recognizable by
Lemma~\ref{lem:tw-finite-recognizable} and
Lemma~\ref{lem:tw-complex-closure}.

We will use one elementary lifting fact.  Let
\[
  I=(i_1,\ldots,i_k)
\]
be a list of distinct coordinates from \(\{1,\ldots,m\}\), and let
\[
  \pi_I:\Sigma_m\to\Sigma_k
\]
be the arity-preserving map that selects these coordinates from each
\(\{0,1\}^m\)-label and sends \(\lambda\) to \(\lambda\).  If
\(R\subseteq(\powfin(N_2))^k\) is recognizable, then the lifted relation
\[
  R^I(\bar A)
  \quad\Longleftrightarrow\quad
  (A_{i_1},\ldots,A_{i_k})\in R
\]
on \(m\)-tuples is recognizable.

Indeed, since \(e(R)\) and \(Z_k\) are recognizable,
\[
  e(R)\cdot_\lambda Z_k
\]
is recognizable by Lemma~\ref{lem:tw-complex-product}.  By inverse projection,
\[
  \pi_I^{-1}(e(R)\cdot_\lambda Z_k)
\]
is recognizable over \(\Sigma_m\), and so is
\[
  L_I=\pi_I^{-1}(e(R)\cdot_\lambda Z_k)\cap U_m.       \tag{1}
\]
For a canonical \(m\)-track term \(t=e(A_1,\ldots,A_m)\), the projection
\(\pi_I(t)\) encodes \((A_{i_1},\ldots,A_{i_k})\), but it may contain redundant
all-zero subtrees created by deleting the other tracks.  Removing exactly those
redundant subtrees gives the canonical encoding
\[
  e(A_{i_1},\ldots,A_{i_k}).
\]
Equivalently,
\[
  \pi_I(t)\in
  \{e(A_{i_1},\ldots,A_{i_k})\}\cdot_\lambda Z_k.       \tag{2}
\]
This equivalence is proved by structural induction on the canonical term \(t\):
at a node whose selected coordinates and selected children are all zero, the
projection contributes an element of \(Z_k\); otherwise one applies the induction
hypothesis to the children.  From (1) and (2) we get
\[
  t\in L_I
  \quad\Longleftrightarrow\quad
  (A_{i_1},\ldots,A_{i_k})\in R.
\]
Thus \(L_I=e(R^I)\), proving the lifting fact.

We now prove the three closure claims.

For conjunction, put all variables occurring in
\[
  R(\bar A,\bar B)\wedge S(\bar B,\bar C)
\]
into one common list.  By the lifting fact, \(R\) and \(S\) can both be viewed as
recognizable relations on this common list.  Their conjunction is then
recognized by intersecting the two lifted recognizable sets, using
Theorem~\ref{thm:tw-boolean}.

For negation, if \(R\subseteq(\powfin(N_2))^n\), then the complement relation is
taken only over canonical encodings:
\[
  e(\neg R)
  =
  U_n\cap\bigl(\Term{\Sigma_n}\setminus e(R)\bigr).
\]
For \(n>0\), the right hand side is recognizable by
Lemma~\ref{lem:tw-encoding-n} and Boolean closure.  For \(n=0\), the canonical
set is \(U_0=\{\lambda\}\), so the same conclusion is immediate from finite
recognizability and Boolean closure.

For existential quantification, it is enough to consider the first coordinate.
Any other coordinate can be moved there by a permutation of tracks; such a
permutation is an arity-preserving bijection of species, so recognizability is
transported by Theorems~\ref{thm:tw-projection} and
\ref{thm:tw-inverse-projection}.  Let
\[
  R(A,\bar B)
  \subseteq
  \powfin(N_2)\times(\powfin(N_2))^{m-1}
\]
be recognizable, and let
\[
  \pi:\Sigma_m\to\Sigma_{m-1}
\]
delete the first coordinate of each label.  By Theorem~\ref{thm:tw-projection},
\[
  \pi(e(R))
\]
is recognizable.  Define
\[
  L=
  \bigl(\pi(e(R))/_{\lambda}Z_{m-1}\bigr)\cap U_{m-1}.       \tag{3}
\]
This set is recognizable by Lemma~\ref{lem:tw-truncation} and, if
\(m-1>0\), Lemma~\ref{lem:tw-encoding-n}.  If \(m-1=0\), then
\[
  U_0=\{\lambda\}
\]
is finite, so the same conclusion follows from
Lemma~\ref{lem:tw-finite-recognizable}.

We claim that \(L=e(\exists A\,R(A,\bar B))\).  First suppose
\[
  \bar B\in \exists A\,R(A,\bar B).
\]
Choose \(A\) with \(R(A,\bar B)\), and set
\[
  t=e(A,\bar B).
\]
Then \(t\in e(R)\), so \(\pi(t)\in\pi(e(R))\).  The term \(\pi(t)\) encodes
\(\bar B\), but it may have all-zero subtrees left over from the deleted
coordinate.  As in (2),
\[
  \pi(t)\in\{e(\bar B)\}\cdot_\lambda Z_{m-1}.
\]
Therefore \(e(\bar B)\in \pi(e(R))/_{\lambda}Z_{m-1}\), and since
\(e(\bar B)\in U_{m-1}\), we get \(e(\bar B)\in L\).

Conversely, suppose \(q=e(\bar B)\in L\).  By the definition of truncation,
there is a term
\[
  r\in \pi(e(R))
  \quad\text{with}\quad
  r\in\{q\}\cdot_\lambda Z_{m-1}.
\]
Choose \(t\in e(R)\) with \(\pi(t)=r\).  Since \(t\in e(R)\), there are
\(A\) and \(\bar B'\) such that
\[
  t=e(A,\bar B')
  \qquad\text{and}\qquad
  R(A,\bar B').
\]
The equality \(r=\pi(t)\) implies that \(r\) encodes \(\bar B'\).  On the other
hand, \(r\in\{q\}\cdot_\lambda Z_{m-1}\), and inserting all-zero terms does not
change the encoded tuple, so \(r\) also encodes \(\bar B\).  Hence
\(\bar B'=\bar B\).  Thus \(R(A,\bar B)\) for some \(A\), proving
\[
  \bar B\in\exists A\,R(A,\bar B).
\]
This proves the equality in (3), and hence existential quantification preserves
recognizability.
\end{proof}

\begin{proof}[Proof of Theorem~\ref{thm:tw-definable-recognizable}]
Let \(\varphi(A_1,\ldots,A_n)\) be a formula of the set-only language, and let
\[
  R_\varphi
  =
  \{(B_1,\ldots,B_n)\in(\powfin(N_2))^n:
      (B_1,\ldots,B_n)\models\varphi\}.
\]
We prove, by structural induction on \(\varphi\), that \(R_\varphi\) is
recognizable.

For atomic formulas, the set-only language has the relation
\[
  A\subseteq B
\]
and the singleton-successor relations
\[
  \widehat R_i(A,B) \qquad (i=1,2).
\]
These are recognizable by Lemma~\ref{lem:tw-atomic}.  Equality of set variables,
if used as primitive notation, is handled as
\[
  A=B \quad\Longleftrightarrow\quad A\subseteq B\wedge B\subseteq A.
\]

Assume the induction hypothesis for the immediate subformulas.  If
\[
  \varphi=\psi\wedge\theta,
\]
then \(R_\psi\) and \(R_\theta\) are recognizable, after viewing them over the
common list of free variables of \(\varphi\).  Lemma~\ref{lem:tw-logical-closure}
gives recognizability of their intersection, which is exactly \(R_\varphi\).

If
\[
  \varphi=\neg\psi,
\]
then \(R_\psi\) is recognizable by induction, and
Lemma~\ref{lem:tw-logical-closure} gives recognizability of its complement
inside the canonical encodings \(U_n\).  This complement is \(R_\varphi\).

Finally, if
\[
  \varphi=\exists A_i\,\psi(A_1,\ldots,A_i,\ldots,A_n),
\]
then \(R_\psi\) is recognizable by induction.  The existentially quantified
relation is the projection of \(R_\psi\) away from the \(i\)-th coordinate, with
the result recanonicalized; Lemma~\ref{lem:tw-logical-closure} proves that this
operation preserves recognizability.  Hence \(R_\varphi\) is recognizable.

The Boolean connective \(\vee\) and universal quantification may either be
treated as abbreviations, or derived from the cases above.  Thus every formula
defines a recognizable relation.

The construction is effective: Lemma~\ref{lem:tw-atomic} gives explicit base
automata or regular descriptions, and every step in
Lemma~\ref{lem:tw-logical-closure} uses one of the effective closure
constructions from Sections 2 and 3.
\end{proof}

\begin{corollary}[Effective MSO-to-automata theorem over \(N_2\)]
For every weak MSO formula over the binary successor tree, one can effectively
construct a finite tree automaton accepting exactly the canonical encodings of
its satisfying finite-set assignments.
\end{corollary}

\begin{proof}
Translate away first-order variables, then apply
Theorem~\ref{thm:tw-definable-recognizable}.  Determinize with
Theorem~\ref{thm:tw-subset} if a deterministic automaton is desired.
\end{proof}

\begin{proof}[Proof of Theorem~\ref{thm:tw-doner}]
We give the proof for \(N_2=\{1,2\}^*\), as \TW{} do.  The same argument works
for \(N_k\): one replaces the binary terms below by \(k\)-ary terms and uses one
successor direction for each letter of the \(k\)-letter alphabet.

The decision procedure has three effective translations.

First, eliminate individual variables.  Given a weak MSO formula
\[
  F(x_1,\ldots,x_r,A_1,\ldots,A_m),
\]
introduce a set variable \(X_{x_j}\) for each individual variable \(x_j\).  This
new set variable is intended to denote the singleton \(\{x_j\}\).  The
set-only language therefore has set variables only, the subset relation
\(\subseteq\), and the singleton-successor predicates \(\widehat R_i\), where
\[
  \widehat R_i(A,B)
  \quad\Longleftrightarrow\quad
  A=\{w\},\ B=\{wi\}
  \text{ for some }w\in N_2.
\]
Using the definable predicate \(\mathrm{Sing}\), the translation is:
\[
  (x=y)^*
  =
  X_x\subseteq X_y\wedge X_y\subseteq X_x
    \wedge \mathrm{Sing}(X_x)\wedge\mathrm{Sing}(X_y),
\]
\[
  (x\in A)^*
  =
  X_x\subseteq A\wedge \mathrm{Sing}(X_x),
\]
\[
  (R_i(x,y))^*
  =
  \widehat R_i(X_x,X_y).
\]
It commutes with Boolean connectives, and quantifiers are translated by
\[
  (\exists x\,G)^*
  =
  \exists X_x\,(\mathrm{Sing}(X_x)\wedge G^*),
  \qquad
  (\exists A\,G)^*
  =
  \exists A\,G^*.
\]
The verification in Section~\ref{sec:tw-eliminate-first-order} proves, by
structural induction on formulas, that this translation preserves truth when
each replacement variable \(X_x\) is interpreted as the singleton \(\{x\}\).
Thus every sentence \(S\) of the original language has an effectively computable
set-only sentence \(S^*\) with the same truth value.

Second, encode finite-set assignments as terms.  For \(n\) free set variables,
use the species \(\Sigma_n\) with one nullary symbol \(\lambda\) and binary
symbols \(\{0,1\}^n\).  A term over \(\Sigma_n\) is read as an \(n\)-track
characteristic function on a finite prefix-closed part of \(N_2\).  The map
\[
  c_n:\Term{\Sigma_n}\to(\powfin(N_2))^n
\]
is defined trackwise, and each tuple \(\bar A\) has a canonical representative
\[
  e(\bar A)
\]
obtained by deleting redundant all-zero subtrees.  Lemmas
\ref{lem:tw-encoding-one} and \ref{lem:tw-encoding-n} show that the set
\[
  U_n=\{e(\bar A):\bar A\in(\powfin(N_2))^n\}
\]
of canonical representatives is recognizable.  The proof of those lemmas is
also effective: it describes \(U_n\) by regular operations, and Theorem
\ref{thm:tw-recognizable-regular} turns this description into a finite
automaton.

Third, compile set-only formulas to automata.  If
\(\varphi(A_1,\ldots,A_n)\) is a set-only formula, let
\[
  R_\varphi
  =
  \{\bar A\in(\powfin(N_2))^n:\bar A\models\varphi\}.
\]
Theorem~\ref{thm:tw-definable-recognizable} proves, by induction on
\(\varphi\), that \(R_\varphi\) is recognizable, meaning that
\[
  e(R_\varphi)=\{e(\bar A):\bar A\models\varphi\}
\]
is a recognizable subset of \(\Term{\Sigma_n}\).  The induction is effective:
atomic formulas are handled by Lemma~\ref{lem:tw-atomic}; conjunction,
negation, and existential set quantification are handled by
Lemma~\ref{lem:tw-logical-closure}, using the effective closure constructions
for Boolean operations, projection, inverse projection, and truncation.

We can now decide a sentence.  Given \(S\), compute the equivalent set-only
sentence \(S^*\).  Apply the effective construction above to \(S^*\), regarded
as a formula with no free variables.  The zero-track alphabet has \(\lambda\)
and one binary all-zero label, and its canonical part is just
\[
  U_0=\{\lambda\}.
\]
Thus the automaton constructed for \(S^*\) recognizes either \(\{\lambda\}\), if
\(S^*\) is true, or \(\emptyset\), if \(S^*\) is false.  Theorem
\ref{thm:tw-emptiness} decides which case holds.

This is the same emptiness test described explicitly by \TW{}.  If the final
set-only sentence is written in the form
\[
  \exists A\,T(A)
  \qquad\text{or}\qquad
  \forall A\,T(A),
\]
construct an automaton \(\Aut=(A,\alpha)\) and final set \(A_F\) recognizing
\[
  e(\{B:T(B)\})\subseteq U_1.
\]
If the sentence is existential, it is true exactly when
\[
  \bh_\Aut(A_F)\ne\emptyset.
\]
If the sentence is universal, it is true exactly when there is no canonical
encoding of a counterexample:
\[
  U_1\cap\bigl(\Term{\Sigma_1}\setminus\bh_\Aut(A_F)\bigr)=\emptyset.
\]
The latter set is recognizable by Lemma~\ref{lem:tw-encoding-one} and Boolean
closure, and Theorem~\ref{thm:tw-emptiness} decides the required emptiness
question.  Therefore the truth of \(S\) is decidable.
\end{proof}

\section{Lean-Oriented Decomposition}

For the Courcelle formalization, this paper suggests the following modules or
proof blocks.
\begin{enumerate}[leftmargin=*]
  \item Ranked signatures and finite terms.
  \item Deterministic bottom-up tree automata and runs.
  \item Nondeterministic automata and subset construction.
  \item Boolean closure, projection, and inverse projection constructions.
  \item Atomic automata for labelled tree vocabularies.
  \item Formula induction compiling MSO formulas to automata.
  \item Optional: canonical-encoding machinery from \TW{} \S4 if we choose to
    formalize the \(N_2\)-subset version literally.
\end{enumerate}

The Courcelle proof itself can use a simpler finite-labelled-tree version of
\TW{}'s Section~4: assignments to MSO variables are represented as extra tracks
on the input tree, and existential quantification erases a track by projection.
In that setting, the truncation machinery is unnecessary unless we introduce a
separate canonicalization step.

\begin{thebibliography}{9}

\bibitem{TW}
James W. Thatcher and Jesse B. Wright.
\newblock Generalized finite automata theory with an application to a decision
problem of second-order logic.
\newblock \emph{Mathematical Systems Theory}, 2(1):57--81, 1968.

\end{thebibliography}

\end{document}
