\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{enumitem}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\newcommand{\MSO}{\ensuremath{\mathrm{MSO}}}
\newcommand{\MSOtwo}{\ensuremath{\mathrm{MSO}_2}}

\title{Toward a Self-Contained Tree-Automaton Theorem\\
  \large Components for \texttt{lecture\_note\_expanded.tex}, keyed to
  Thatcher--Wright (1968)}
\author{}
\date{}

\begin{document}
\maketitle

\section*{Purpose and setting}

The main note isolates the effective \MSO{}--tree-automaton correspondence as a
cited theorem.  Neither \textsc{mathlib} nor \textsc{cslib} currently provides a
tree-automata API, so a self-contained (and eventually formalised) development
must build it from scratch.  This file lists the required components and, for each
one, points to the corresponding result in Thatcher and Wright~\cite{TW}, whose
\S\,2--3 develop finite tree-automata theory and whose \S\,4 carries the induction
from logic to automata.

Two remarks keep the citations honest.
\begin{enumerate}[leftmargin=*,label=(\arabic*)]
  \item Only one direction is needed for Courcelle's algorithm: from an \MSO{}
    sentence over the tree vocabulary, compute a \emph{deterministic} bottom-up
    tree automaton recognising its models, and run it.  The converse
    (recognisable \(\Rightarrow\) definable) is not required.
  \item Reference~\cite{TW} proves decidability of the weak monadic second-order
    theory of \(k\) successors (Theorem~14, due to Doner): its ``trees'' are terms
    \(T_\Sigma\) of a dyadic species that \emph{encode finite subsets of}
    \(N_k\).  The note instead works with finite \emph{labelled} binary trees,
    which are already terms of a dyadic species.  Hence the finite-tree-automata
    core (\S\,2--3) and the logic-to-automata induction (\S\,4) transfer directly,
    while the \(N_k\)-subset encoding of \S\,4 (the maps \(c\), \(e\) and
    Lemmas~15--16) is replaced by reading the labelled tree itself.
\end{enumerate}

\section{Objects and definitions}
\begin{itemize}[leftmargin=*]
  \item \emph{Species} \((\Sigma,\sigma)\); a \(\Sigma\)-algebra
    \(\mathcal A=(A,\alpha)\), i.e.\ a deterministic bottom-up tree automaton; the
    term set \(T_\Sigma\) (finite \(\Sigma\)-labelled trees); the run/homomorphism
    \(h_{\mathcal A}\); the behaviour \(\mathrm{bh}_{\mathcal A}(A_F)\); and the
    notion of a \emph{recognisable} set.
    \hfill\emph{Source:} \cite[\S\,2, pp.~59--61]{TW}.
  \item Nondeterministic tree automata (relational algebras) and their runs.
    \hfill\emph{Source:} \cite[\S\,2, pp.~60--61]{TW}.
  \item Encoding an assignment of \(n\) free set variables as extra label tracks:
    the product species \(\Sigma^{n}\) with \(\Sigma^{n}_2=\{0,1\}^n\), one
    \(\{0,1\}\)-track per set variable.
    \hfill\emph{Source:} \cite[\S\,4, step~(II) p.~74; Lemma~16 p.~78]{TW}.
\end{itemize}

\section{Base automata (atomic formulas)}
A tree automaton for each atomic constraint over the track alphabet: membership
\(x\in X\), the label predicates \(P_a\), the child relations
\(\operatorname{child}_1,\operatorname{child}_2\) (the successor relations
\(R_\sigma\)), and equality.  First-order variables are eliminated by treating
them as singleton set variables through a definable \(\mathrm{Sing}\) predicate,
so that \(x\in X\) and \(x=y\) become inclusion/singleton conditions on tracks.
\hfill\emph{Source:} translation and \(\mathrm{Sing}\), \cite[\S\,4,
pp.~74--75]{TW}; recognisability of the atomic relations, \cite[Lemma~18,
pp.~78--79]{TW} (via regular sets and Theorem~8).

\section{Closure operations (the technical core)}
\begin{itemize}[leftmargin=*]
  \item Conjunction and disjunction by the product automaton.
    \hfill\emph{Source:} \cite[Theorem~2(ii), p.~62]{TW}.
  \item \textbf{Determinisation} by the bottom-up subset construction (equivalence
    of nondeterministic and deterministic tree automata), with the finite
    powerset state space.
    \hfill\emph{Source:} \cite[Theorem~1, pp.~61--62]{TW}.
  \item Negation by complementing a \emph{deterministic} automaton
    (\(A_F\mapsto A\setminus A_F\)); thus \(\neg\) depends on determinisation.
    \hfill\emph{Source:} \cite[Theorem~2(i), p.~62]{TW}.
  \item Existential set quantification \(\exists X\) by projection (erasing one
    track), which yields a nondeterministic automaton.
    \hfill\emph{Source:} \cite[Theorem~3, pp.~62--63]{TW}; the dual inverse
    projection (adding a track) is \cite[Theorem~4, pp.~63--64]{TW}.
\end{itemize}

\section{Compilation theorem (definable \(\Rightarrow\) recognisable)}
By induction on an \MSO{} formula with \(n\) free variables, build an automaton
over \(\Sigma^{n}\) recognising exactly the (tree, assignment) pairs that satisfy
it: atomic formulas by the base automata; \(\wedge\) and \(\neg\) by Theorem~2;
\(\exists\) by the projection Theorem~3.  Determinise the resulting automaton by
Theorem~1.
\hfill\emph{Source:} \cite[Theorem~17, p.~78]{TW} (using Lemma~18 and
Theorems~1--3).

\section{Faithfulness of the tree encoding}
Match the note's ordered, labelled binary trees to terms of a dyadic species, so
that \(\operatorname{child}_1,\operatorname{child}_2\) and the label predicates
are read off the term structure.  In \cite{TW} this is the correspondence between
dyadic-species terms and finite (prefix-closed) subsets of \(N_2\): the
characteristic map \(c\), the canonical encoding \(e\) (the smallest term), and
the recognisability of the set of encodings.
\hfill\emph{Source:} \cite[\S\,4, pp.~75--78; Lemmas~15--16]{TW}.  For finite
labelled trees this step is light, since the tree is already the term.

\section{Execution, emptiness, and complexity}
Acceptance is decided by computing the run \(h_{\mathcal A}\) bottom-up.  Emptiness
of the behaviour is decidable---needed for the decision procedure, though not for
model-checking a single fixed tree---via the replacement lemma and a depth bound.
\hfill\emph{Source:} run \(h_{\mathcal A}\), \cite[\S\,2, pp.~60--61]{TW};
replacement, \cite[Lemma~5, p.~65; Lemma~6, p.~65]{TW}; emptiness,
\cite[Theorem~7, pp.~65--66]{TW}.  The \emph{linear-time} bound on a single
bottom-up pass of a deterministic automaton is not addressed in \cite{TW} and
remains a metatheoretic claim requiring a cost model.

\section{Optional: the regularity characterisation}
Reference~\cite{TW} also proves ``recognisable \(\Leftrightarrow\) regular'' (a
Kleene-type theorem for trees), used there to exhibit the atomic automata.  It is
convenient but not strictly required if the atomic automata are constructed
directly.
\hfill\emph{Source:} \cite[Theorem~8, pp.~66--67; Theorem~9, pp.~67--69;
Lemmas~10--13, pp.~69--72]{TW}.

\section{Scope remark}
The development is sizeable---essentially the finite-tree form of the
B\"uchi--Elgot--Trakhtenbrot theorem, namely \S\,2--3 of \cite{TW} together with
the \S\,4 induction.  Until it is carried out, the cited correspondence should be
regarded as the single substantial external dependency of the note.

\begin{thebibliography}{9}

\bibitem{TW}
James W. Thatcher and Jesse B. Wright.
\newblock Generalized finite automata theory with an application to a decision
problem of second-order logic.
\newblock \emph{Mathematical Systems Theory}, 2(1):57--81, 1968.

\end{thebibliography}

\end{document}
